\providecommand{\Nb}{\mathcal{N}}
\providecommand{\GL}{\operatorname{GL}}
\providecommand{\Aut}{\operatorname{Aut}}
\providecommand{\tr}{\operatorname{tr}}
\providecommand{\HASH}{\operatorname{HASH}}
\providecommand{\deg}{\operatorname{deg}}

\chapter{Geometric Priors: Invariance and Equivariance}
\label{ch:priors}

The previous chapters introduced the mathematical language of geometric deep
learning (GDL) --- groups, representations, and the geometric domains on which
data live --- and argued that the failure of classical architectures on
non-Euclidean data is conceptual rather than technical. This chapter turns that
diagnosis into a constructive principle. The central thesis is simple to state:
a model should respect the symmetries of the problem it solves, and the way to
guarantee this is to build the symmetry into the architecture rather than to
hope it will be learned from data. The two notions that make this precise are
\emph{invariance} and \emph{equivariance}. From them we derive a design
blueprint --- the \emph{geometric priors} of the title --- that organises the
domains and symmetries of GDL around five recurring themes and unifies
convolutional, attention-based, and graph architectures under a single
geometric construction: message passing.

\section{The geometric deep learning paradigm}
\label{sec:priors:paradigm}

Geometric deep learning extends deep learning beyond the traditional Euclidean
domains --- regular grids of pixels, linear sequences of tokens --- towards
arbitrary geometric structures. Molecules are three-dimensional graphs with
rotational symmetries; social networks are irregular graphs with emergent
topological properties; physical surfaces are Riemannian manifolds with
intrinsic curvature. GDL provides the mathematical framework needed to build
neural architectures that respect and exploit these natural geometries instead
of restricting attention to representations on Euclidean domains.

One of the foundational contributions of GDL is the demonstration that the
seemingly disparate architectures of deep learning --- convolutional neural
networks (CNNs), transformers, graph networks --- are particular instances of a
single unifying principle: the equivariant processing of signals over geometric
domains with specific symmetries. CNNs emerge as the case of regular grids with
translational invariance; transformers as complete graphs in which attention
implements a weighted message passing; and graph neural networks (GNNs) as the
general case of arbitrary graphs with permutation invariance. This unifying
perspective rests on deep mathematical tools from group theory, differential
geometry, and harmonic analysis, providing not only theoretical foundations for
why existing architectures work, but also design principles for new ones with
guarantees of stability and generalization.

\subsection{The five fundamental Gs}
\label{subsec:priors:five-gs}

The GDL framework is structured around five fundamental
concepts, the ``five Gs''~\citep{bronstein2017geometric,
bronstein2021geometric}, which provide a unified vocabulary for understanding
and designing neural architectures that exploit the geometry where the data
live.

\begin{description}
\item[Geometry.] The intrinsic structure of the data domain and its metric
properties: Riemannian manifolds (such as spheres for omnidirectional data),
hyperbolic spaces (for hierarchical data), and product spaces (for
multi-structured data). The geometry determines which operations are natural and
efficient on each domain.
\item[Groups.] The symmetries that preserve the structure of the problem under
transformation. The most common are translations $\R^d$ (CNNs), rotations
$\mathrm{SO}(n)$ (rotation-equivariant networks), permutations $S_n$ (graph
networks), and Lorentz groups (relativistic networks). Identifying the correct
symmetry group is crucial for architectural design.
\item[Graphs.] Discretizations of continuous geometric domains that capture
local neighbourhood relations. Graphs may be regular (uniform meshes), irregular
(social networks), directed (causal networks), or dynamic (evolving systems).
They are the computational substrate for many GDL architectures.
\item[Geodesics.] Optimal paths that respect the intrinsic geometry of the
space. On Riemannian manifolds geodesics minimize distance; on graphs they
correspond to shortest paths; on information spaces they define directions of
steepest gradient. Geodesics determine how information propagates in geometric
architectures.
\item[Gauges.] Invariances under gauge transformations (local changes of
reference frame) that preserve the observable physics of the system. These
include electromagnetic invariances, coordinate freedoms on manifolds, and
orientation choices on undirected graphs. Gauges ensure that the learned
representations are intrinsic to the geometric domain.
\end{description}

\Cref{fig:priors:five-gs} arranges the five Gs around the central GDL paradigm,
and \Cref{fig:priors:hierarchy} complements this picture by exhibiting their
hierarchical structure, from the most fundamental layer (the geometry of the
domain) to the most specific level of abstraction (local frames).

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    node distance=1.8cm,
    gnode/.style={
        circle, draw=gdlgray!80, fill=#1, minimum size=2.0cm,
        font=\bfseries\small
    },
    arrow/.style={-{Stealth[length=8pt]}, thick, draw=gdlgray!60},
    lbl/.style={font=\small\itshape, text=gdlgray!70}
]
\node[gnode=gdlgray!25, minimum size=2.4cm] (gdl) at (0,0) {GDL};
\node[gnode=gdlblue!20] (geometry) at (90:3.6cm) {Geometry};
\node[gnode=gdlgreen!20] (groups) at (162:3.6cm) {Groups};
\node[gnode=gdlorange!20] (graphs) at (234:3.6cm) {Graphs};
\node[gnode=gdlred!20] (geodesics) at (306:3.6cm) {Geodesics};
\node[gnode=gdlpurple!20] (gauges) at (18:3.6cm) {Gauges};
\draw[arrow] (gdl) -- (geometry);
\draw[arrow] (gdl) -- (groups);
\draw[arrow] (gdl) -- (graphs);
\draw[arrow] (gdl) -- (geodesics);
\draw[arrow] (gdl) -- (gauges);
\node[lbl, above=0.05cm of geometry] {domains};
\node[lbl, left=0.05cm of groups] {symmetries};
\node[lbl, below left=0.05cm of graphs] {discretization};
\node[lbl, below right=0.05cm of geodesics] {distances};
\node[lbl, right=0.05cm of gauges] {coordinates};
\draw[thick, dashed, draw=gdlgray!40] (geometry) to[bend left=15] (groups);
\draw[thick, dashed, draw=gdlgray!40] (groups) to[bend left=15] (graphs);
\draw[thick, dashed, draw=gdlgray!40] (graphs) to[bend left=15] (geodesics);
\draw[thick, dashed, draw=gdlgray!40] (geodesics) to[bend left=15] (gauges);
\draw[thick, dashed, draw=gdlgray!40] (gauges) to[bend left=15] (geometry);
\end{tikzpicture}
\caption[The five Gs of geometric deep learning]{The five fundamental Gs of
geometric deep learning. Each component represents an essential aspect of the
unifying framework: \textbf{Geometry} (domains), \textbf{Groups} (symmetries),
\textbf{Graphs} (discretization), \textbf{Geodesics} (distances), and
\textbf{Gauges} (local coordinates).}
\label{fig:priors:five-gs}
\end{figure}

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=0.8,
    level/.style={rectangle, rounded corners=3pt, draw, minimum height=0.7cm,
        align=center, font=\small\bfseries},
    arrow/.style={->, thick, >=stealth, gray!70},
    desc/.style={font=\scriptsize, text=black!60, align=left}
]
\node[level, fill=gdlgreen!15, minimum width=10cm] (geom) at (0,0) {Geometry};
\node[level, fill=gdlgreen!25, minimum width=8cm] (grupos) at (0,1.2) {Groups};
\node[level, fill=gdlgreen!35, minimum width=6cm] (grafos) at (0,2.4) {Graphs};
\node[level, fill=gdlgreen!45, minimum width=4cm] (geod) at (0,3.6) {Geodesics};
\node[level, fill=gdlgreen!55, minimum width=2.5cm] (gauges) at (0,4.8) {Gauges};
\draw[arrow] (geom.north) ++(0,0.1) -- ++(0,0.3);
\draw[arrow] (grupos.north) ++(0,0.1) -- ++(0,0.3);
\draw[arrow] (grafos.north) ++(0,0.1) -- ++(0,0.3);
\draw[arrow] (geod.north) ++(0,0.1) -- ++(0,0.3);
\node[desc, anchor=west] at (geom.east) {\quad domain $\Omega$ where the data live};
\node[desc, anchor=west] at (grupos.east) {\quad symmetries $\group$ that preserve structure};
\node[desc, anchor=west] at (grafos.east) {\quad discretization of the domain};
\node[desc, anchor=west] at (geod.east) {\quad information propagation};
\node[desc, anchor=west] at (gauges.east) {\quad local invariances};
\node[font=\scriptsize, anchor=east] at (geom.west) {$\R^n$, $\mathcal{M}$, $\mathcal{G}$\quad};
\node[font=\scriptsize, anchor=east] at (grupos.west) {$\mathrm{SO}(n)$, $S_n$, $E(n)$\quad};
\node[font=\scriptsize, anchor=east] at (grafos.west) {$\mathcal{G}=(V,E)$\quad};
\node[font=\scriptsize, anchor=east] at (geod.west) {$d(x,y)$, paths\quad};
\node[font=\scriptsize, anchor=east] at (gauges.west) {frames\quad};
\end{tikzpicture}
\caption[Hierarchy of the five Gs]{Hierarchical structure of the five Gs of
geometric deep learning. From the base, the \textbf{Geometry} defines the
domain where the data reside; the \textbf{Groups} capture the symmetries of the
problem; the \textbf{Graphs} discretize the domain; the \textbf{Geodesics}
determine how information propagates; and the \textbf{Gauges} handle local
invariances in coordinate systems.}
\label{fig:priors:hierarchy}
\end{figure}

From this unifying perspective, GDL holds that the effective processing of
signals over geometric domains requires respecting three fundamental principles.

\begin{description}
\item[Geometric locality.] Operations should be local with respect to the metric
of the domain. Geometric architectures exploit local structure through
neighbourhoods defined by the underlying geometry, letting information propagate
consistently with the intrinsic distances of the space (formalised through
graphs in \Cref{ch:graphs}).
\item[Equivariance.] Transformations should preserve the relevant symmetries of
the problem. An equivariant function commutes with the actions of the symmetry
group, guaranteeing that equivalent transformations of the input produce
corresponding transformations of the output
(\Cref{sec:priors:inv-equi}).
\item[Compositionality.] Complex representations are built hierarchically by
composing simpler operations. The composition of equivariant layers preserves
equivariance end to end (\Cref{thm:priors:composition}), enabling deep
architectures that retain the desired geometric properties. This property
underpins the equivariant universal approximation theorem
(\Cref{thm:priors:universal}), which guarantees the expressive power of
equivariant networks.
\end{description}

\section{Why traditional networks need a geometric prior}
\label{sec:priors:motivation}

A traditional artificial neural network treats its input as an unstructured
vector $x \in \R^n$. A single neuron computes $y = \sigma(\inner{w}{x} + b)$,
combining all $n$ coordinates through a global weighted sum. Reinterpreted as an
operator on a signal over a domain $\Omega = \{1, \dots, n\}$, this domain is a
bare index set: it carries no notion of \emph{locality} (every coordinate
contributes equally, with no concept of near and far), no \emph{structured
neighbourhood} (the connection $w_1 x_1$ is structurally identical to
$w_{100} x_{100}$), and no \emph{geometric invariance} (permuting the indices
generally changes the output, even when the permutation is semantically
irrelevant). The network ignores any geometry implicit in the data.

This blindness has a concrete cost. Suppose the quantity we wish to predict is
unchanged under a group $\group$ of transformations --- rotations of an image,
relabellings of the nodes of a graph. A fully connected network has no built-in
reason to produce the same answer for $x$ and for a transformed input
$g \cdot x$. The standard remedy, \emph{data augmentation}, presents the network
with transformed copies of every example; but this inflates the training set by
a factor of $\abs{\group}$, offers no guarantee that the learned function is
actually invariant, and degrades or fails entirely when $\group$ is large or
continuous. A geometric prior replaces this hope with a guarantee: the symmetry
is encoded in the structure of the map, so it holds exactly, for every input,
by construction.

\section{Invariant and equivariant functions}
\label{sec:priors:inv-equi}

The notions of equivariance and invariance form the theoretical core of GDL.
Having developed the language of groups and representations (\Cref{ch:foundations}),
we are now in a position to formalise precisely how a function can respect the
symmetries of a problem. We work with a group $\group$ acting on the relevant
spaces. Recall that an \emph{action} of $\group$ on a set $X$ is a map
$\group \times X \to X$, written $(g, x) \mapsto g \cdot x$, that is compatible
with the group law: $e \cdot x = x$ and $g \cdot (h \cdot x) = (gh) \cdot x$.
When $X$ is a vector space and each $g$ acts linearly, the action is a
\emph{representation} $\rho_X \colon \group \to \GL(X)$. We write $g \cdot x$
for $\rho_X(g)\, x$ throughout.

\begin{definition}[Equivariant function]
\label{def:priors:equivariance}
Let $\group$ be a group acting on $X$ via $\rho_X$ and on $Y$ via $\rho_Y$.
A map $\phi \colon X \to Y$ is \emph{$\group$-equivariant} if
\[
  \phi(g \cdot x) = g \cdot \phi(x)
  \qquad \text{for all } g \in \group,\ x \in X,
\]
where $g \cdot x$ denotes $\rho_X(g)\, x$ and $g \cdot \phi(x)$ denotes
$\rho_Y(g)\, \phi(x)$; equivalently, $\phi \circ \rho_X(g) = \rho_Y(g) \circ \phi$
for every $g$.
\end{definition}

\begin{definition}[Invariant function]
\label{def:priors:invariance}
A map $\phi \colon X \to Y$ is \emph{$\group$-invariant} if
$\phi(g \cdot x) = \phi(x)$ for all $g \in \group$ and $x \in X$. This is
equivalent to $\group$-equivariance when the action of $\group$ on $Y$ is
trivial, $\rho_Y(g) = \mathrm{id}$.
\end{definition}

Invariance is thus the special case of equivariance in which $\group$ acts
trivially on the output. Equivariance is the more fundamental notion: it says
that transforming the input produces a correspondingly transformed output, so
that geometric structure is \emph{preserved} rather than discarded. In many GDL
applications we seek equivariant intermediate layers that preserve structure,
followed by a final invariant layer.

\Cref{fig:priors:equivariance} renders \Cref{def:priors:equivariance} as a
commutative square: equivariance is exactly the statement that the two paths
from $X$ to $Y$ --- transform then apply $\phi$, or apply $\phi$ then transform
--- agree.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    node distance=2.6cm and 4cm,
    space/.style={circle, draw=gdlgray!80, fill=gdlblue!15, minimum size=1.5cm, font=\large},
    arrow/.style={-{Stealth[length=7pt]}, thick},
    lbl/.style={font=\small, midway}
]
\node[space] (X) {$X$};
\node[space, right=of X] (Y) {$Y$};
\node[space, below=of X] (gX) {$X$};
\node[space, below=of Y] (gY) {$Y$};
\draw[arrow, gdlblue!70!black] (X) -- node[lbl, above] {$\phi$} (Y);
\draw[arrow, gdlblue!70!black] (gX) -- node[lbl, below] {$\phi$} (gY);
\draw[arrow, gdlred!70!black] (X) -- node[left, align=center] {\small\itshape transform\\[1pt] $\rho_X(g)$} (gX);
\draw[arrow, gdlred!70!black] (Y) -- node[lbl, right] {$\rho_Y(g)$} (gY);
\node[font=\Large] at ($(X)!0.5!(gY)$) {$\circlearrowleft$};
\node[font=\normalsize, below=0.4cm of gX, anchor=north west, xshift=0.6cm] {
    $\phi(\rho_X(g)\cdot x) = \rho_Y(g)\cdot \phi(x)$
};
\end{tikzpicture}
\caption[The equivariance commutative square]{Equivariance as a commutative
diagram. A map $\phi \colon X \to Y$ is $\group$-equivariant when the square
commutes: transforming the input by $\rho_X(g)$ and then applying $\phi$ yields
the same result as applying $\phi$ and then transforming the output by
$\rho_Y(g)$.}
\label{fig:priors:equivariance}
\end{figure}

Traditional artificial neural networks lack equivariance by design. To
approximate invariance they require \emph{data augmentation} with all the
transformations of the group $\group$, inflating the training set by a factor of
$\abs{\group}$ with no guarantee of generalization. GDL resolves this by building
the symmetries directly into the architecture, propagating equivariance through
all layers.

\begin{example}[Three canonical symmetries]
\label{ex:priors:symmetries}
Image classification is invariant to translations: shifting the picture of a cat
does not change the label, so the score function should be $\Z^2$-invariant.
Semantic segmentation, by contrast, is \emph{equivariant} to translations:
shifting the image shifts the segmentation mask by the same amount. A function
defined on the nodes of a graph should be equivariant to the permutation group
$S_n$: relabelling the nodes permutes the outputs identically, because a graph
carries no canonical ordering of its vertices. Each task fixes a group $\group$
and a required behaviour --- invariant or equivariant --- under it.
\end{example}

The reason equivariance is the right primitive for deep architectures is that it
is preserved by composition.

\begin{theorem}[Composition of equivariant maps]
\label{thm:priors:composition}
Let $\group$ act on $X$, $Y$, $Z$ via representations $\rho_X, \rho_Y, \rho_Z$.
If $f \colon X \to Y$ and $h \colon Y \to Z$ are $\group$-equivariant, then so
is the composition $h \circ f \colon X \to Z$.
\end{theorem}

\begin{proof}
Fix $g \in \group$ and $x \in X$. Using $\group$-equivariance of $f$ and then of
$h$,
\[
  (h \circ f)(g \cdot x) = h\bigl(f(g \cdot x)\bigr)
    = h\bigl(g \cdot f(x)\bigr)
    = g \cdot h\bigl(f(x)\bigr)
    = g \cdot (h \circ f)(x). \qedhere
\]
\end{proof}

\Cref{thm:priors:composition} is the keystone of the whole framework. It
guarantees that a network built by stacking equivariant layers is itself
equivariant, end to end --- equivariance is a property we establish once, per
layer, and then inherit for free at every depth. Composing such an equivariant
backbone with a single invariant readout at the end yields a globally invariant
predictor. This turns symmetry from a property to be checked \emph{a posteriori}
into a constructive recipe for entire architectures.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    >=Stealth,
    layer/.style={rectangle, draw, thick, rounded corners, minimum width=1.9cm, minimum height=1.1cm, font=\small},
    data/.style={circle, draw, thick, minimum size=0.95cm, font=\small},
    arrow/.style={->, thick}
]
\node[data, fill=gdlblue!20] (x) at (0, 0) {$x$};
\node[layer, fill=gdlgreen!20] (f1) at (2.8, 0) {Layer $f_1$};
\node[layer, fill=gdlgreen!20] (f2) at (5.6, 0) {Layer $f_2$};
\node[layer, fill=gdlgreen!20] (f3) at (8.4, 0) {Layer $f_3$};
\node[data, fill=gdlred!20] (y) at (11.2, 0) {$y$};
\draw[arrow] (x) -- (f1);
\draw[arrow] (f1) -- (f2);
\draw[arrow] (f2) -- (f3);
\draw[arrow] (f3) -- (y);
\node[data, fill=gdlblue!40] (gx) at (0, -2.8) {$g \cdot x$};
\node[data, fill=gdlred!40] (gy) at (11.2, -2.8) {$g \cdot y$};
\draw[arrow, gdlpurple!70] (x) -- node[left, font=\small] {$\rho_X(g)$} (gx);
\draw[arrow, gdlpurple!70] (y) -- node[right, font=\small] {$\rho_Y(g)$} (gy);
\draw[arrow, dashed, gdlorange!70] (gx) -- (8.7, -2.8) -- (gy);
\node[font=\small, gdlpurple!70] at (1.4, -1.4) {transform};
\node[font=\small, gdlorange!70] at (5.6, -3.3) {$(f_3 \circ f_2 \circ f_1)(g \cdot x)$};
\node[font=\Large] at (11.2, -1.4) {$=$};
\node[font=\tiny, above] at (2.8, 0.65) {$\rho_1$-equiv};
\node[font=\tiny, above] at (5.6, 0.65) {$\rho_2$-equiv};
\node[font=\tiny, above] at (8.4, 0.65) {$\rho_3$-equiv};
\end{tikzpicture}
\caption[Composition of equivariant layers]{Composition of equivariant layers.
If each layer $f_i$ is $\group$-equivariant, then the composition
$f_3 \circ f_2 \circ f_1$ is $\group$-equivariant: transforming the input and
then applying the network yields the same result as applying the network and
then transforming the output.}
\label{fig:priors:composition}
\end{figure}

\section{The geometric deep learning blueprint}
\label{sec:priors:blueprint}

The composition principle suggests a template for designing architectures on any
geometric domain~\citep{bronstein2021geometric}. Fix a domain $\Omega$ (a grid,
a graph, a manifold), a space of signals $\mathcal{X}(\Omega)$ on it, and a
symmetry group $\group$ acting on both. The \emph{geometric deep learning
blueprint} assembles a model from four kinds of building block.

\begin{description}
\item[Equivariant layers.] Linear or nonlinear maps
$\mathcal{X}(\Omega) \to \mathcal{X}(\Omega')$ that commute with the action of
$\group$. These are the workhorses: they extract features while transforming
predictably under the symmetry. A pointwise nonlinearity applied after an
equivariant linear map preserves equivariance as long as it acts identically on
every position.
\item[Local pooling (coarsening).] Equivariant maps that reduce the resolution
of the domain, $\Omega \to \Omega'$, while respecting the symmetry. Pooling
implements \emph{scale separation}: it lets later layers see larger
neighbourhoods through a coarser domain, building a hierarchy of features.
\item[Invariant global readout.] A final $\group$-invariant map that collapses
the signal to the desired output --- a single label, a graph-level property ---
typically a symmetric aggregation (sum, mean, max) followed by a standard
network.
\item[Symmetry-respecting attention or message functions.] When connectivity
itself should adapt to content, the weights linking positions are computed by
functions that are themselves equivariant. This generalises fixed local filters
to data-dependent ones.
\end{description}

Two structural principles underlie this template and recur throughout the book.

\begin{description}
\item[Symmetry (equivariance).] Every internal operation commutes with
$\group$, so the global function is equivariant by
\Cref{thm:priors:composition}, with invariance achieved only at the readout.
\item[Scale separation (locality and compositionality).] Operations are
\emph{local} with respect to the geometry of $\Omega$ --- each layer aggregates
only over a neighbourhood --- and complex representations are built
\emph{compositionally}, by stacking local layers interleaved with coarsening.
Locality keeps the parameter count and computation tractable; compositionality
lets a deep stack of local operations capture long-range, global structure.
\end{description}

\begin{remark}[One blueprint, five domains]
\label{rem:priors:fivegs}
The blueprint is deliberately domain-agnostic. Instantiating it on a grid with
the translation group recovers convolutional networks (\Cref{ch:grids});
extending the group to rotations and reflections gives group-equivariant
networks (\Cref{ch:groups}); replacing the grid by an arbitrary graph with the
permutation group gives graph neural networks (\Cref{ch:graphs}); endowing the
domain with intrinsic distances gives learning on manifolds
(\Cref{ch:geodesics}); and acknowledging the absence of a canonical local frame
on a curved domain gives gauge-equivariant networks (\Cref{ch:gauges}). The
five Gs are five concrete realisations of one geometric idea.
\end{remark}

\section{Universality and expressivity}
\label{sec:priors:universality}

A natural worry is that constraining a network to be equivariant might cripple
its expressive power: the constraint carves out a strict subspace of all
functions. The following result, which extends the classical universal
approximation theorem to the equivariant setting, dispels that worry. It is due
to \citet{yarotsky2022} and \citet{kondor2018generalizationconvolutionalneural}.

\begin{theorem}[Equivariant universal approximation]
\label{thm:priors:universal}
Let $\group$ be a compact group acting continuously on compact metric spaces $X$
and $Y$, and let $\mathcal{F}^{\group}(X, Y)$ be the space of all continuous
$\group$-equivariant maps $f \colon X \to Y$. Then for every
$f \in \mathcal{F}^{\group}(X, Y)$ and every $\varepsilon > 0$ there exists a
$\group$-equivariant neural network $\mathcal{N}$ such that
\[
  \sup_{x \in X} d_Y\bigl(f(x), \mathcal{N}(x)\bigr) < \varepsilon,
\]
where $d_Y$ is the metric on $Y$.
\end{theorem}

\begin{proof}
The argument proceeds in four steps, following the ideas of
\citet{yarotsky2022}.

\emph{Step 1: non-equivariant approximation.} By the classical universal
approximation theorem, for any continuous $f \colon X \to Y$ and any
$\varepsilon > 0$ there is an unconstrained neural network
$\tilde f \colon X \to Y$ with
$\sup_{x \in X} d_Y(f(x), \tilde f(x)) < \varepsilon$.

\emph{Step 2: equivariant averaging.} Define the \emph{Reynolds operator} by
averaging over the group with respect to the normalized Haar measure $\mu$ on
$\group$:
\[
  \bar f(x) = \int_{\group} \rho_Y(g^{-1})\, \tilde f(g \cdot x)\, d\mu(g),
\]
where $\rho_Y \colon \group \to \GL(Y)$ is the representation of $\group$ on $Y$.

\emph{Step 3: verification of equivariance.} We check that $\bar f$ is
$\group$-equivariant. For any $h \in \group$,
\[
  \bar f(h \cdot x)
    = \int_{\group} \rho_Y(g^{-1})\, \tilde f(g \cdot (h \cdot x))\, d\mu(g)
    = \int_{\group} \rho_Y(g^{-1})\, \tilde f((gh) \cdot x)\, d\mu(g).
\]
Changing variables to $g' = gh$ and using the invariance of the Haar measure
($d\mu(g) = d\mu(g')$),
\[
  \bar f(h \cdot x)
    = \int_{\group} \rho_Y\bigl((g'h^{-1})^{-1}\bigr)\, \tilde f(g' \cdot x)\, d\mu(g')
    = \rho_Y(h) \int_{\group} \rho_Y(g'^{-1})\, \tilde f(g' \cdot x)\, d\mu(g')
    = \rho_Y(h)\, \bar f(x).
\]
Hence $\bar f$ is $\group$-equivariant.

\emph{Step 4: control of the approximation error.} Let $f$ be the target
$\group$-equivariant function. By linearity of the integral and equivariance of
$f$,
\[
  d_Y\bigl(\bar f(x), f(x)\bigr)
    \le \int_{\group} d_Y\bigl(\rho_Y(g^{-1}) \tilde f(g \cdot x),\,
        \rho_Y(g^{-1}) f(g \cdot x)\bigr)\, d\mu(g).
\]
For a compact group the representations may be taken unitary, so $\rho_Y(g^{-1})$
preserves distances and
\[
  d_Y\bigl(\bar f(x), f(x)\bigr)
    \le \int_{\group} d_Y\bigl(\tilde f(g \cdot x), f(g \cdot x)\bigr)\, d\mu(g)
    \le \sup_{x' \in X} d_Y\bigl(\tilde f(x'), f(x')\bigr) < \varepsilon,
\]
the last step using that the action of $\group$ on the compact $X$ preserves
compactness. Approximating $\bar f$ by a $\group$-equivariant network completes
the proof. \qedhere
\end{proof}

The message is decisive: \emph{equivariance costs nothing in expressivity}. A
$\group$-equivariant network can approximate any continuous $\group$-equivariant
target to arbitrary accuracy. The symmetry constraint does not shrink the set of
reachable functions within the relevant subspace; it acts purely as an inductive
bias, discarding the irrelevant non-equivariant functions and so improving
sample efficiency and generalization without sacrificing capacity.

\begin{example}[CNNs as a special case]
\label{ex:priors:cnn-universal}
For the translation group $\group = \Z^2$ acting on discrete images, the theorem
guarantees that CNNs --- which are translation-equivariant
(\Cref{ch:grids}) --- can approximate any continuous function invariant or
equivariant to translations. This justifies theoretically the empirical success
of convolutional networks in computer vision~\citep{cohen2016group}.
\end{example}

\section{Message passing: the unifying framework}
\label{sec:priors:message-passing}

The five Gs form a deeply interconnected conceptual framework: the
\emph{geometry} determines the structure of the domain and which \emph{groups}
of symmetries are natural; \emph{graphs} discretize the continuous geometries
while preserving topological and spectral properties; \emph{geodesics} define
optimal paths that respect the geometry; and \emph{gauges} ensure that
representations are intrinsic. From this perspective the differences between
neural architectures reflect choices about three geometric aspects: the
\emph{structure of the domain} (Euclidean spaces, graphs, manifolds), the
\emph{symmetry group} (trivial, translations, permutations), and the
\emph{connectivity} (global, fixed neighbourhoods, adaptive). Traditional
artificial neural networks, operating in $\R^d$ with trivial symmetries,
constitute the base case: they exploit no geometric structure, so all invariance
must be learned from data. GDL resolves this by building the symmetries directly
into the architecture through the principle of equivariance.

The blueprint becomes algorithmically concrete through a single computational
primitive: \emph{message passing}. Viewing data as a graph
$\mathcal{G} = (V, E)$ of nodes and edges, a layer propagates information by
having each node aggregate transformed features from its neighbours. This one
construction subsumes the major deep-learning architectures; their apparent
differences reduce to choices of graph topology, connectivity, and symmetry
group.

\begin{definition}[Message-passing neural network]
\label{def:priors:mpnn}
A \emph{message-passing neural network} (MPNN) operates on a graph
$\mathcal{G} = (V, E)$ with node features $\{h_v\}_{v \in V}$ and edge features
$\{e_{uv}\}_{(u,v) \in E}$. The update at layer $\ell$ consists of two phases,
a \emph{message-passing} phase and a \emph{node-update} phase:
\[
  m_v^{(\ell+1)} = \bigoplus_{u \in \Nb(v)}
      M_\ell\bigl(h_v^{(\ell)}, h_u^{(\ell)}, e_{vu}\bigr),
  \qquad
  h_v^{(\ell+1)} = U_\ell\bigl(h_v^{(\ell)}, m_v^{(\ell+1)}\bigr),
\]
where $\Nb(v)$ is the neighbourhood of $v$, $M_\ell$ a learnable message
function, $\bigoplus$ a permutation-invariant aggregator (sum, mean, max), and
$U_\ell$ an update function~\citep{gilmer2017neural}. This scheme operates
exclusively over nodes and edges; its generalization to higher-order
$k$-simplices, \emph{simplicial message passing}, is developed in
\Cref{sec:priors:tdl}.
\end{definition}

The aggregator $\bigoplus$ is permutation-invariant by design, which makes each
layer equivariant to relabellings of the neighbours --- the message-passing
layer is the blueprint's equivariant layer for relational data. The following
correspondences (\Cref{fig:priors:mpnn}) show that the familiar architectures
are special cases.

\begin{figure}[h]
\centering
\begin{tikzpicture}[>=Stealth, font=\small,
    nd/.style={circle, draw, thick, fill=white, minimum size=4mm, inner sep=0pt},
    edge/.style={thick, draw=gdlgray!60},
    box/.style={rectangle, rounded corners=4pt, draw=gdlgray!70, thick, minimum width=4.6cm, minimum height=3.7cm},
    title/.style={font=\small\bfseries, fill=white, inner sep=2pt},
    info/.style={font=\scriptsize, text=black!70, align=center}
]
% ANN
\begin{scope}[shift={(-3.3, 2.7)}]
    \node[box, fill=gdlblue!15] (annbox) {};
    \node[title] at (0, 2.05) {ANN (fully connected)};
    \foreach \i/\a in {1/90,2/162,3/234,4/306,5/18} \node[nd, fill=gdlblue!20] (a\i) at (\a:0.9) {};
    \foreach \i in {1,...,5} \foreach \j in {1,...,5} {\ifnum\i<\j \draw[edge, gdlblue!30] (a\i)--(a\j);\fi}
    \node[info] at (0, -1.45) {complete graph $K_d$\\specific weights $W_{ij}$};
\end{scope}
% CNN
\begin{scope}[shift={(3.3, 2.7)}]
    \node[box, fill=gdlgreen!15] (cnnbox) {};
    \node[title] at (0, 2.05) {CNN (convolution)};
    \foreach \i in {0,1,2} \foreach \j in {0,1,2} \node[nd, fill=gdlgreen!20] (c\i\j) at ({\i*0.7-0.7},{\j*0.7-0.7}) {};
    \foreach \i in {0,1,2} \foreach \j [evaluate=\j as \jn using int(\j+1)] in {0,1} {
        \draw[edge, gdlgreen!50] (c\i\j)--(c\i\jn); \draw[edge, gdlgreen!50] (c\j\i)--(c\jn\i);}
    \draw[thick, gdlorange, rounded corners=2pt] (-0.3,-0.3) rectangle (0.3,0.3);
    \node[info] at (0, -1.45) {grid $\Z^d$, shared weights\\$\group=(\Z^d,+)$};
\end{scope}
% Transformer
\begin{scope}[shift={(-3.3, -2.7)}]
    \node[box, fill=gdlpurple!15] (transbox) {};
    \node[title] at (0, 2.05) {Transformer (attention)};
    \foreach \i/\a in {1/90,2/162,3/234,4/306,5/18} \node[nd, fill=gdlpurple!20] (t\i) at (\a:0.9) {};
    \draw[edge, gdlpurple!60, line width=2pt] (t1)--(t3);
    \draw[edge, gdlpurple!40, line width=1.4pt] (t1)--(t2);
    \draw[edge, gdlpurple!20, line width=0.5pt] (t1)--(t4);
    \draw[edge, gdlpurple!50, line width=1.4pt] (t2)--(t4);
    \draw[edge, gdlpurple!60, line width=2pt] (t3)--(t5);
    \draw[edge, gdlpurple!30, line width=0.8pt] (t4)--(t5);
    \node[info] at (0, -1.45) {complete graph $K_n$\\adaptive weights $\alpha_{ij}$};
\end{scope}
% GNN
\begin{scope}[shift={(3.3, -2.7)}]
    \node[box, fill=gdlred!15] (gnnbox) {};
    \node[title] at (0, 2.05) {GNN (graph network)};
    \node[nd, fill=gdlred!20] (g1) at (-0.7,0.6) {};
    \node[nd, fill=gdlred!20] (g2) at (0.3,0.8) {};
    \node[nd, fill=gdlred!20] (g3) at (0.8,0.1) {};
    \node[nd, fill=gdlred!20] (g4) at (-0.4,-0.3) {};
    \node[nd, fill=gdlred!20] (g5) at (0.4,-0.5) {};
    \node[nd, fill=gdlred!20] (g6) at (-0.15,0.25) {};
    \foreach \a/\b in {g1/g2,g1/g4,g1/g6,g2/g3,g2/g6,g3/g5,g4/g5,g4/g6,g5/g6}
        \draw[edge, gdlred!50] (\a)--(\b);
    \node[info] at (0, -1.45) {arbitrary graph $\mathcal{G}$\\$\group=\Aut(\mathcal{G})$};
\end{scope}
% Centre
\node[font=\bfseries, align=center, fill=gdlyellow!20, rounded corners=3pt,
      draw=gdlyellow!70, thick, inner sep=5pt] (mpnn) at (0,0)
      {MPNN\\[-1pt]{\scriptsize $m_v=\bigoplus_{u\in\Nb(v)} M(h_v,h_u,e_{vu})$}};
\draw[->, thick, gdlblue!60] (annbox.south east) -- (mpnn.north west);
\draw[->, thick, gdlgreen!50!black] (cnnbox.south west) -- (mpnn.north east);
\draw[->, thick, gdlpurple!60] (transbox.north east) -- (mpnn.south west);
\draw[->, thick, gdlred!60] (gnnbox.north west) -- (mpnn.south east);
\end{tikzpicture}
\caption[Architectures as message passing]{The major deep-learning
architectures as instances of message passing. They differ only in graph
topology, weight-sharing scheme, and symmetry group $\group$: fully connected
networks (complete graph, pair-specific weights), convolutional networks (grid,
translation-shared weights), transformers (complete graph, content-adaptive
weights), and graph networks (arbitrary graph). The message-passing update at
the centre is common to all.}
\label{fig:priors:mpnn}
\end{figure}

\subsection{Fully connected layers as trivial MPNNs}
\label{subsec:priors:ann-mpnn}

\begin{proposition}[ANN as an MPNN on a uniform complete graph]
\label{prop:priors:ann-mpnn}
A fully connected layer $h^{(\ell+1)} = \sigma(W h^{(\ell)} + b)$ is equivalent
to an MPNN where the graph is the complete graph $K_d$ whose features are nodes,
the message is the pair-specific $M(h_v, h_u, e_{vu}) = W_{vu} h_u$, aggregation
is the sum over all input features, and the update is
$U(h_v, m_v) = \sigma(m_v + b_v)$.
\end{proposition}

\begin{proof}
Let $h^{(\ell)} \in \R^d$ be the feature vector at layer $\ell$. The fully
connected layer computes
\[
  h_v^{(\ell+1)} = \sigma\!\left(\sum_{u=1}^{d} W_{vu}\, h_u^{(\ell)} + b_v\right).
\]
Interpreting each component as a node of the complete graph $K_d$, this
expression coincides exactly with the message passing of
\Cref{def:priors:mpnn}, with message $M(h_v, h_u, e_{vu}) = W_{vu} h_u^{(\ell)}$,
aggregation $m_v = \sum_{u \in \Nb(v)} M(\cdot) = \sum_{u=1}^d W_{vu} h_u^{(\ell)}$,
and update $U(h_v, m_v) = \sigma(m_v + b_v)$. \qedhere
\end{proof}

This reveals why fully connected networks lack a \emph{geometric inductive bias}:
they treat every connection between features as equally important, ignoring any
spatial, topological, or similarity structure the data may possess.

\subsection{Convolutions as regular message passing}
\label{subsec:priors:cnn-mpnn}

CNNs are the first non-trivial instance of message passing, in which the grid
structure imposes a fundamental geometric inductive bias: translation
equivariance.

\begin{proposition}[CNN as an MPNN on a regular grid]
\label{prop:priors:cnn-mpnn}
A convolutional layer on a regular grid $\Z^d$ with kernel support $\mathcal{K}$
is equivalent to an MPNN with neighbourhood
$\Nb(v) = \{u \in \Z^d : v - u \in \mathcal{K}\}$, message
$M(h_v, h_u, e_{vu}) = W_{v-u}\, h_u$ (weights shared by relative position),
sum aggregation over the local neighbourhood, and update
$U(h_v, m_v) = \sigma(m_v + b)$.
\end{proposition}

\begin{proof}
Let $h^{(\ell)} \colon \Z^d \to \R^{c_{\mathrm{in}}}$ be the signal at layer
$\ell$ and $\{W_\delta\}_{\delta \in \mathcal{K}}$ the convolutional kernel with
$W_\delta \in \R^{c_{\mathrm{out}} \times c_{\mathrm{in}}}$. The discrete
convolution computes
\[
  h_v^{(\ell+1)}
    = \sigma\!\left(\sum_{\delta \in \mathcal{K}} W_\delta\, h_{v-\delta}^{(\ell)} + b\right)
    = \sigma\!\left(\sum_{u \in \Nb(v)} W_{v-u}\, h_u^{(\ell)} + b\right),
\]
where the change of variable $u = v - \delta$ turns the sum over kernel
displacements into a sum over the neighbours of $v$ in the grid graph. This
matches the message passing of \Cref{def:priors:mpnn} with message
$M(h_v, h_u, e_{vu}) = W_{v-u} h_u^{(\ell)}$, aggregation
$m_v = \sum_{u \in \Nb(v)} M(\cdot)$, and update $U(h_v, m_v) = \sigma(m_v + b)$.
\qedhere
\end{proof}

The key difference from fully connected networks is that the weights $W_{v-u}$
depend only on the relative position, not on the absolute pair $(v, u)$. This
\emph{weight sharing} is precisely the implementation of translation
equivariance as a geometric inductive bias, reducing the parameter count
drastically from $O(n^2 c^2)$ to $O(\abs{\mathcal{K}}\, c^2)$.

\subsection{Attention as adaptive message passing}
\label{subsec:priors:attention-mpnn}

Transformers represent the complementary case: the communication weights are
computed dynamically from the content of the nodes, not from their position.

\begin{proposition}[Transformer as an MPNN on an adaptive complete graph]
\label{prop:priors:attention-mpnn}
A self-attention layer on a sequence of $n$ tokens is equivalent to an MPNN on
the complete graph $K_n$ whose nodes are the tokens, with message
$M(h_i, h_j) = \alpha_{ij}\, V h_j$ and attention coefficients
\[
  \alpha_{ij}
    = \frac{\exp\!\bigl((Q h_i)^{\!\top}(K h_j)/\sqrt{d_k}\bigr)}
           {\sum_{m=1}^{n} \exp\!\bigl((Q h_i)^{\!\top}(K h_m)/\sqrt{d_k}\bigr)},
\]
weighted-sum aggregation $m_i = \sum_{j=1}^{n} M(h_i, h_j)$, and the identity
update $U(h_i, m_i) = m_i$.
\end{proposition}

\begin{proof}
Let $\{h_1, \dots, h_n\} \subset \R^d$ be the input sequence and
$Q, K, V \in \R^{d \times d_k}$ the projection matrices. The self-attention
layer computes
\[
  h_i^{(\ell+1)} = \sum_{j=1}^{n} \alpha_{ij}\, V h_j^{(\ell)}
    = \sum_{j=1}^{n} \underbrace{\alpha_{ij}\, V h_j^{(\ell)}}_{M(h_i, h_j)}.
\]
Identifying the complete graph $K_n$ with $\Nb(i) = \{1, \dots, n\}$, this is
exactly the message passing of \Cref{def:priors:mpnn} with message
$M(h_i, h_j) = \alpha_{ij} V h_j$ and identity update $U(h_i, m_i) = m_i$. The
softmax normalization guarantees $\alpha_{ij} \in (0, 1)$ with
$\sum_{j=1}^{n} \alpha_{ij} = 1$, so the aggregation is a convex combination,
unlike the uniform sum of fully connected layers
(\Cref{prop:priors:ann-mpnn}) and convolutions
(\Cref{prop:priors:cnn-mpnn})~\citep{vaswani2017attentionneed}. \qedhere
\end{proof}

The fundamental difference from the previous architectures is that the weights
$\alpha_{ij}$ are functions of the content $(h_i, h_j)$, neither fixed (fully
connected) nor dependent only on relative position (convolutions). This
adaptivity allows global connectivity at the cost of quadratic complexity
$O(n^2)$ in the sequence length.

Reading across these cases reveals a single axis of increasing geometric
sophistication: fully connected networks impose no prior (uniform, pair-specific
weights); convolutions impose a \emph{spatial} prior (weights shared by relative
position, encoding translation equivariance); transformers replace fixed weights
by \emph{adaptive, content-dependent} ones (global connectivity at quadratic
cost); and graph networks generalise the whole scheme to arbitrary relational
topologies, equivariant to the automorphism group $\Aut(\mathcal{G})$. The
geometric prior is precisely what each architecture assumes about which
connections matter and how their weights are shared.

\subsection{Topological deep learning as higher-order message passing}
\label{sec:priors:tdl}

The three architectures above (ANNs, CNNs, transformers) operate exclusively
over the \emph{nodes} ($0$-simplices) of a graph, propagating information along
edges. Topological deep learning (TDL) breaks this restriction: it assigns
signals to $k$-simplices of arbitrary order and propagates information
simultaneously through the lower and upper adjacencies defined by the boundary
operators. This extension incorporates a \emph{topological inductive bias} that
captures higher-order relations inaccessible to standard message passing.

\begin{definition}[Simplicial signal]
\label{def:priors:simplicial-signal}
A \emph{simplicial signal of order $k$} on a simplicial complex $\mathcal{K}$ is
a function $\mathbf{x} \colon \mathcal{K}_k \to \R^d$ assigning a feature vector
$\mathbf{x}_\sigma \in \R^d$ to each $k$-simplex $\sigma \in \mathcal{K}_k$.
\end{definition}

\begin{definition}[Simplicial message passing]
\label{def:priors:simplicial-mp}
\emph{Simplicial message passing} generalises message passing on graphs
(\Cref{def:priors:mpnn}) through three kinds of communication for each
$k$-simplex $\sigma$:
\begin{enumerate}
\item \textbf{Lower message} (via $\partial_k^{\!\top}$): information from the
$(k{-}1)$-simplices on the boundary of $\sigma$,
\[
  \mathbf{m}_\sigma^{\mathrm{down}}
    = \bigoplus_{\tau \in \mathcal{B}^{\mathrm{down}}(\sigma)}
      M^{\mathrm{down}}\bigl(\mathbf{h}_\sigma, \mathbf{h}_\tau\bigr),
\]
where $\mathcal{B}^{\mathrm{down}}(\sigma) =
\{\tau : [\partial_k^{\!\top} \partial_k]_{\sigma,\tau} \neq 0\}$ are the
$k$-simplices adjacent through shared faces.
\item \textbf{Upper message} (via $\partial_{k+1}$): information from the
$(k{+}1)$-simplices that contain $\sigma$,
\[
  \mathbf{m}_\sigma^{\mathrm{up}}
    = \bigoplus_{\rho \in \mathcal{B}^{\mathrm{up}}(\sigma)}
      M^{\mathrm{up}}\bigl(\mathbf{h}_\sigma, \mathbf{h}_\rho\bigr),
\]
where $\mathcal{B}^{\mathrm{up}}(\sigma) =
\{\rho : [\partial_{k+1} \partial_{k+1}^{\!\top}]_{\sigma,\rho} \neq 0\}$ are the
$k$-simplices sharing a $(k{+}1)$-simplex.
\item \textbf{Update}:
$\mathbf{h}_\sigma' = U\bigl(\mathbf{h}_\sigma,
\mathbf{m}_\sigma^{\mathrm{down}}, \mathbf{m}_\sigma^{\mathrm{up}}\bigr)$.
\end{enumerate}
Here $\bigoplus$ denotes a permutation-invariant aggregation (sum, mean, max),
$M^{\mathrm{down}}, M^{\mathrm{up}}$ are learnable message functions, and $U$ is
the update function~\citep{bodnar2021weisfeiler}.
\end{definition}

\begin{proposition}[TDL as a generalized instance of the MPNN framework]
\label{prop:priors:tdl-mpnn}
Simplicial message passing (\Cref{def:priors:simplicial-mp}) is a generalized
instance of the MPNN framework (\Cref{def:priors:mpnn}) with domain the
simplicial complex $\mathcal{K}$, the $k$-simplices $\sigma \in \mathcal{K}_k$ as
``nodes''; double neighbourhood
$\Nb(\sigma) = \mathcal{B}^{\mathrm{down}}(\sigma) \cup
\mathcal{B}^{\mathrm{up}}(\sigma)$ determined by the lower and upper Laplacians
$L_k^{\mathrm{down}} = \partial_k^{\!\top} \partial_k$ and
$L_k^{\mathrm{up}} = \partial_{k+1} \partial_{k+1}^{\!\top}$; the combination of
lower and upper messages; and the learnable update
$U(\mathbf{h}_\sigma, \mathbf{m}_\sigma^{\mathrm{down}},
\mathbf{m}_\sigma^{\mathrm{up}})$. In particular, for $k = 0$ one recovers the
standard MPNN on graphs.
\end{proposition}

\begin{proof}
We verify the correspondence with the three components of the MPNN framework.
Fix $k \ge 0$ and consider the graph $\mathcal{G}_k = (\mathcal{K}_k, E_k)$ whose
nodes are the $k$-simplices and whose edges encode the adjacency given by $L_k$.

\emph{(1) Message.} For each $k$-simplex $\sigma$, simplicial message passing
computes
\[
  \mathbf{m}_\sigma^{\mathrm{down}}
    = \bigoplus_{\tau \in \mathcal{B}^{\mathrm{down}}(\sigma)}
      M^{\mathrm{down}}(\mathbf{h}_\sigma, \mathbf{h}_\tau),
  \qquad
  \mathbf{m}_\sigma^{\mathrm{up}}
    = \bigoplus_{\rho \in \mathcal{B}^{\mathrm{up}}(\sigma)}
      M^{\mathrm{up}}(\mathbf{h}_\sigma, \mathbf{h}_\rho).
\]
Defining $\Nb(\sigma) = \mathcal{B}^{\mathrm{down}}(\sigma) \cup
\mathcal{B}^{\mathrm{up}}(\sigma)$ and the joint message
$\widetilde{M}(\mathbf{h}_\sigma, \mathbf{h}_\nu) =
M^{\mathrm{down}}(\mathbf{h}_\sigma, \mathbf{h}_\nu)$ when
$\nu \in \mathcal{B}^{\mathrm{down}}(\sigma)$ and
$\widetilde{M}(\mathbf{h}_\sigma, \mathbf{h}_\nu) =
M^{\mathrm{up}}(\mathbf{h}_\sigma, \mathbf{h}_\nu)$ when
$\nu \in \mathcal{B}^{\mathrm{up}}(\sigma)$, this takes the form
$m_\sigma = \bigoplus_{\nu \in \Nb(\sigma)} \widetilde{M}(\mathbf{h}_\sigma,
\mathbf{h}_\nu)$, which is exactly the message phase of \Cref{def:priors:mpnn}.

\emph{(2) Aggregation.} The operator $\bigoplus$ (sum, mean, or max) plays the
role of the permutation-invariant aggregation of the MPNN scheme.

\emph{(3) Update.} The function
$U(\mathbf{h}_\sigma, \mathbf{m}_\sigma^{\mathrm{down}},
\mathbf{m}_\sigma^{\mathrm{up}})$ corresponds directly to the update function
$U_\ell$ of the MPNN framework.

\emph{Reduction to $k = 0$.} For $0$-simplices (nodes), $\partial_0 = 0$ gives
$L_0^{\mathrm{down}} = 0$, so $\mathcal{B}^{\mathrm{down}}(v) = \emptyset$ and the
lower message vanishes. The upper message reduces to
$\mathbf{m}_v^{\mathrm{up}} = \bigoplus_{u \in \mathcal{B}^{\mathrm{up}}(v)}
M^{\mathrm{up}}(\mathbf{h}_v, \mathbf{h}_u)$, where
$\mathcal{B}^{\mathrm{up}}(v) = \{u : [\partial_1 \partial_1^{\!\top}]_{vu} \neq 0\}
= \Nb(v)$ is the standard neighbourhood of $v$ in the $1$-skeleton. This recovers
exactly the classical message passing on graphs. \qedhere
\end{proof}

\begin{proposition}[Expressivity of simplicial message passing]
\label{prop:priors:simplicial-expressivity}
Simplicial message passing is strictly more expressive than message passing on
graphs. There exist pairs of non-isomorphic simplicial complexes whose
$1$-skeletons are identical (and hence indistinguishable by MPNNs) but which
simplicial networks distinguish correctly thanks to higher-order topological
information.
\end{proposition}

This result, due to \citet{bodnar2021weisfeiler}, shows that the topological
information encoded in the boundary operators $\partial_k$ provides a
discriminative power that surpasses the limitations of the Weisfeiler--Leman
test established in \Cref{sec:priors:wl}. It closes the unification arc of the
MPNN framework: whereas ANNs, CNNs, and transformers encode spatial or
relational inductive biases over nodes, TDL introduces a \emph{higher-order
topological inductive bias} that captures interactions among triangles,
tetrahedra, and arbitrary $k$-simplices, yielding strictly greater expressive
power and crossing the Weisfeiler--Leman
barrier~\citep{bodnar2021weisfeiler, hajij2023topological}.

\section{Expressive power and its limits}
\label{sec:priors:expressivity-limits}

Message passing is unifying but not omnipotent. Its expressive power on graphs is
pinned down by a classical combinatorial algorithm, and its depth is limited by
two complementary pathologies. We develop these here; the spectral and geometric
analysis returns in \Cref{ch:graphs,ch:geodesics}.

\subsection{The Weisfeiler--Leman test}
\label{sec:priors:wl}

The \emph{Weisfeiler--Leman test} (1-WL) is a classical algorithm, originally
proposed in 1968~\citep{weisfeiler1968reduction} to decide whether two graphs are
isomorphic through an iterative process of \emph{colour refinement}. Its
importance in GDL is that it establishes a fundamental upper bound on the
expressive power of MPNNs.

\begin{definition}[Weisfeiler--Leman algorithm (1-WL)]
\label{def:priors:wl}
Let $\mathcal{G} = (V, E)$ be a graph with $n$ vertices. The
\emph{Weisfeiler--Leman algorithm of order 1} (1-WL) consists of the following
iterative colouring process.
\begin{enumerate}
\item \textbf{Initialization} ($t = 0$): assign each node $v \in V$ an initial
colour $c^{(0)}(v)$, typically identical for all nodes (or based on node labels
if present).
\item \textbf{Iterative refinement}: at each iteration $t \ge 0$, update the
colour of each node by
\[
  c^{(t+1)}(v) = \HASH\Bigl(c^{(t)}(v),\,
      \{\!\{ c^{(t)}(u) : u \in \Nb(v) \}\!\}\Bigr),
\]
where $\Nb(v)$ is the neighbourhood of $v$, $\{\!\{\cdot\}\!\}$ denotes a
multiset, and $\HASH$ is an injective function assigning a new unique colour to
each distinct pair (old colour, multiset of neighbour colours).
\item \textbf{Convergence}: the algorithm terminates when the colour partition
stabilizes, i.e.\ when $c^{(t+1)}(v) = c^{(t)}(v)$ for all $v \in V$. This occurs
in at most $n$ iterations.
\item \textbf{Final histogram}: the result is the \emph{colour histogram}
$\mathbf{h}(\mathcal{G}) = (|\{v : c^{(T)}(v) = k\}|)_k$, counting how many nodes
carry each final colour.
\end{enumerate}
\end{definition}

\begin{definition}[WL-equivalence]
\label{def:priors:wl-equivalence}
Two graphs $\mathcal{G}_1$ and $\mathcal{G}_2$ are \emph{WL-equivalent}, written
$\mathcal{G}_1 \equiv_{\mathrm{WL}} \mathcal{G}_2$, if the 1-WL algorithm produces
the same colour histogram for both.
\end{definition}

\begin{figure}[h]
\centering
\begin{tikzpicture}[>=Stealth, font=\small,
    wlnode/.style={circle, draw, thick, minimum size=7mm, inner sep=0pt},
    stage/.style={font=\bfseries\small}
]
\newcommand{\drawedges}{%
    \draw[thick, gdlgray!50] (n1) -- (n2);
    \draw[thick, gdlgray!50] (n2) -- (n3);
    \draw[thick, gdlgray!50] (n3) -- (n4);
    \draw[thick, gdlgray!50] (n4) -- (n1);
    \draw[thick, gdlgray!50] (n3) -- (n5);
    \draw[thick, gdlgray!50] (n4) -- (n5);
    \draw[thick, gdlgray!50] (n1) -- (n6);
    \draw[thick, gdlgray!50] (n2) -- (n6);
    \draw[thick, gdlgray!50] (n3) -- (n6);
    \draw[thick, gdlgray!50] (n4) -- (n6);
}
% (a) initial
\begin{scope}[shift={(0,0)}]
    \node[stage] at (1, 3) {(a) Initial: $c^{(0)}$};
    \node[wlnode, fill=gdlgray] (n1) at (0, 1.8) {};
    \node[wlnode, fill=gdlgray] (n2) at (2, 1.8) {};
    \node[wlnode, fill=gdlgray] (n3) at (2, 0) {};
    \node[wlnode, fill=gdlgray] (n4) at (0, 0) {};
    \node[wlnode, fill=gdlgray] (n5) at (1, -1.2) {};
    \node[wlnode, fill=gdlgray] (n6) at (1, 0.9) {};
    \drawedges
    \node[font=\tiny, above left=-1pt] at (n1) {$v_1$};
    \node[font=\tiny, above right=-1pt] at (n2) {$v_2$};
    \node[font=\tiny, right=1pt] at (n3) {$v_3$};
    \node[font=\tiny, left=1pt] at (n4) {$v_4$};
    \node[font=\tiny, below=1pt] at (n5) {$v_5$};
    \node[font=\tiny, above=1pt] at (n6) {$v_6$};
\end{scope}
\draw[->, very thick, gdlgray!50] (3.3, 0.9) -- node[above, font=\scriptsize] {HASH} (4.7, 0.9);
% (b) refinement by degree
\begin{scope}[shift={(5.5,0)}]
    \node[stage] at (1, 3) {(b) Iteration 1: $c^{(1)}$};
    \node[wlnode, fill=gdlblue!50] (n1) at (0, 1.8) {};
    \node[wlnode, fill=gdlblue!50] (n2) at (2, 1.8) {};
    \node[wlnode, fill=gdlred!50] (n3) at (2, 0) {};
    \node[wlnode, fill=gdlred!50] (n4) at (0, 0) {};
    \node[wlnode, fill=gdlgreen!50] (n5) at (1, -1.2) {};
    \node[wlnode, fill=gdlred!50] (n6) at (1, 0.9) {};
    \drawedges
\end{scope}
\draw[->, very thick, gdlgray!50] (8.8, 0.9) -- node[above, font=\scriptsize] {HASH} (10.2, 0.9);
% (c) further refinement
\begin{scope}[shift={(11,0)}]
    \node[stage] at (1, 3) {(c) Iteration 2: $c^{(2)}$};
    \node[wlnode, fill=gdlblue!50] (n1) at (0, 1.8) {};
    \node[wlnode, fill=gdlblue!50] (n2) at (2, 1.8) {};
    \node[wlnode, fill=gdlyellow!60] (n3) at (2, 0) {};
    \node[wlnode, fill=gdlyellow!60] (n4) at (0, 0) {};
    \node[wlnode, fill=gdlgreen!50] (n5) at (1, -1.2) {};
    \node[wlnode, fill=gdlred!50] (n6) at (1, 0.9) {};
    \drawedges
    \node[font=\scriptsize, text=black!50, anchor=west] at (2.3, 0.9) {stable};
\end{scope}
\node[anchor=west, font=\scriptsize] at (0, -2.6) {%
    Refinement: $c^{(t+1)}(v) = \mathrm{HASH}\!\big(c^{(t)}(v),\, \{\!\{c^{(t)}(u) : u \in \mathcal{N}(v)\}\!\}\big)$};
\node[anchor=west, font=\scriptsize, text=black!60] at (0, -3.3) {%
    Each node updates its colour from its current colour and the multiset of its neighbours' colours.};
\node[anchor=west] at (0, -4.2) {%
    \tikz{\node[wlnode, fill=gdlblue!50, minimum size=4mm] {};} {\scriptsize degree 3} \quad
    \tikz{\node[wlnode, fill=gdlred!50, minimum size=4mm] {};} {\scriptsize degree 4 (type A)} \quad
    \tikz{\node[wlnode, fill=gdlyellow!60, minimum size=4mm] {};} {\scriptsize degree 4 (type B)} \quad
    \tikz{\node[wlnode, fill=gdlgreen!50, minimum size=4mm] {};} {\scriptsize degree 2}
};
\end{tikzpicture}
\caption[Weisfeiler--Leman colour refinement]{The 1-WL colour-refinement test on
a six-node graph. (a) Uniform initial colouring. (b) After the first iteration,
nodes are distinguished by degree: blue (degree 3), red (degree 4), green
(degree 2). (c) The second iteration refines according to the multiset of
neighbour colours: $v_3$ and $v_4$ (yellow) separate from $v_6$ (red) because
their neighbourhoods have different colour compositions.}
\label{fig:priors:wl}
\end{figure}

The deep connection between the WL test and graph networks was established
rigorously by \citet{xu2019how} and \citet{morris2019weisfeiler}.

\begin{theorem}[WL--MPNN equivalence]
\label{thm:priors:wl-mpnn}
Let $\mathcal{G}$ be a class of graphs and let $\mathcal{A}$ be an MPNN with
injective aggregation and update functions. Then:
\begin{enumerate}
\item \textbf{Upper bound}: $\mathcal{A}$ cannot distinguish WL-equivalent
graphs. That is, if $\mathcal{G}_1 \equiv_{\mathrm{WL}} \mathcal{G}_2$, then
$\mathcal{A}(\mathcal{G}_1) = \mathcal{A}(\mathcal{G}_2)$.
\item \textbf{Optimality}: there exists an MPNN (specifically, a Graph
Isomorphism Network) that is \emph{as expressive as} the WL test. That is, if
$\mathcal{G}_1 \not\equiv_{\mathrm{WL}} \mathcal{G}_2$, there is a weight
configuration such that
$\mathcal{A}(\mathcal{G}_1) \neq \mathcal{A}(\mathcal{G}_2)$.
\end{enumerate}
\end{theorem}

\begin{proof}[Proof sketch]
The key is the structural correspondence between one iteration of WL and one
layer of an MPNN.

\emph{WL $\to$ MPNN}: the WL refinement
$c^{(t+1)}(v) = \HASH(c^{(t)}(v), \{\!\{c^{(t)}(u)\}\!\})$ maps directly to a
message-aggregate-update layer
\[
  h_v^{(t+1)} = \phi\Bigl(h_v^{(t)},
      \bigoplus_{u \in \Nb(v)} \psi(h_u^{(t)})\Bigr).
\]
If $\phi$ and $\bigoplus$ are sufficiently expressive (e.g.\ a sum over unique
embeddings followed by an MLP), they reproduce the behaviour of $\HASH$ exactly.

\emph{MPNN $\to$ WL}: conversely, if two graphs are WL-equivalent they share the
same sequence of colour partitions. Since an MPNN can only aggregate information
from the neighbourhood, it cannot break the symmetry that the WL test preserves.

The full proof requires showing that the sum aggregator over sufficiently unique
embeddings is injective, which follows from the universal approximation theorem
for sets~\citep{zaheer2017deepsets}. \qedhere
\end{proof}

\begin{example}[Non-isomorphic graphs indistinguishable by WL]
\label{ex:priors:wl-indistinguishable}
The simplest non-isomorphic graphs that 1-WL cannot distinguish are regular
graphs of equal degree and size. Consider:
\begin{itemize}
\item $\mathcal{G}_1$: a six-node cycle (hexagon), $C_6$;
\item $\mathcal{G}_2$: two disjoint triangles, $K_3 \sqcup K_3$.
\end{itemize}
Both graphs have $6$ nodes, $6$ edges, and every node has degree $2$. The WL
algorithm assigns:
\begin{enumerate}
\item Iteration 0: all nodes receive the same initial colour $c^{(0)} = 1$.
\item Iteration 1: each node sees neighbours of colour $1$, multiset
$\{\!\{1, 1\}\!\}$. New colour $c^{(1)} = \HASH(1, \{\!\{1, 1\}\!\}) = 2$ for all.
\item Convergence: the partition does not change. Final histogram
$\mathbf{h}(\mathcal{G}_1) = \mathbf{h}(\mathcal{G}_2) = (6)$.
\end{enumerate}
Yet $\mathcal{G}_1 \not\cong \mathcal{G}_2$: the cycle $C_6$ is connected while
$K_3 \sqcup K_3$ has two connected components. This limitation shows that
standard MPNNs cannot detect the global connectivity of regular graphs. If a task
requires distinguishing molecules with similar global structure but different
connectivity (e.g.\ structural isomers), standard MPNNs may fail
systematically.
\end{example}

\subsection{Oversmoothing: spectral analysis}
\label{sec:priors:oversmoothing}

\Cref{thm:priors:wl-mpnn} delimits which graphs MPNNs can distinguish, but says
nothing about what happens when we stack many message-passing layers. The
answer, known as \emph{oversmoothing}~\citep{li2018deeper}, is that node
representations converge to a uniform value, losing all discriminative power.
This phenomenon admits an elegant spectral explanation.

\begin{definition}[Dirichlet energy on graphs]
\label{def:priors:dirichlet-energy}
Let $\mathcal{G} = (V, E)$ be a graph with Laplacian $L_G$ and let
$H \in \R^{n \times d}$ be a matrix of node features ($n = |V|$, $d$ channels).
The \emph{Dirichlet energy} of $H$ is
\[
  E(H) = \tr(H^{\!\top} L_G H)
       = \sum_{c=1}^{d} \mathbf{h}_c^{\!\top} L_G \mathbf{h}_c
       = \sum_{c=1}^{d} \sum_{(i,j) \in E} W_{ij}\,(h_{ic} - h_{jc})^2,
\]
where $\mathbf{h}_c$ denotes the $c$-th column of $H$.
\end{definition}

The Dirichlet energy measures the total variation of the signals $H$ over the
graph: $E(H) = 0$ if and only if $H$ is constant on each connected component.

\begin{proposition}[Linear MPNN as discrete diffusion]
\label{prop:priors:mpnn-diffusion}
Consider an MPNN (\Cref{def:priors:mpnn}) with mean aggregation and no
nonlinearity,
\[
  h_v^{(\ell+1)} = (1 - \alpha)\, h_v^{(\ell)}
      + \alpha \sum_{u \in \Nb(v)} \frac{h_u^{(\ell)}}{\deg(v)},
  \qquad \alpha \in (0, 1].
\]
Let $\tilde L = D^{-1} L_G = I - D^{-1} A$ be the random-walk Laplacian, which
shares its spectrum with the symmetric normalized Laplacian. Then the update is,
in matrix form,
\[
  H^{(\ell+1)} = (I - \alpha \tilde L)\, H^{(\ell)}.
\]
\end{proposition}

\begin{proof}
For each node $v$ the update rule is
\[
  h_v^{(\ell+1)}
    = (1 - \alpha) h_v^{(\ell)} + \alpha \sum_{u \sim v} \frac{h_u^{(\ell)}}{\deg(v)}
    = h_v^{(\ell)} - \alpha\Bigl(h_v^{(\ell)} - \sum_{u \sim v} \frac{h_u^{(\ell)}}{\deg(v)}\Bigr)
    = h_v^{(\ell)} - \alpha (\tilde L\, H^{(\ell)})_v,
\]
using $\tilde L = D^{-1} L_G = I - D^{-1} A$ and
$(D^{-1} A H)_v = \sum_{u \sim v} h_u / \deg(v)$. \qedhere
\end{proof}

This is the temporal discretization of the heat equation
$\partial_t H = -\alpha \tilde L H$ on the graph. Iterating $\ell$ layers gives
$H^{(\ell)} = (I - \alpha \tilde L)^\ell H^{(0)}$, which diffuses the signals
progressively.

\begin{theorem}[Exponential convergence of oversmoothing]
\label{thm:priors:oversmoothing}
Let $\mathcal{G}$ be a connected graph with normalized-Laplacian eigenvalues
$0 = \lambda_1 < \lambda_2 \le \cdots \le \lambda_n$ and
$\alpha \in (0, 1/\lambda_n)$. Then the Dirichlet energy of the representations
after $\ell$ layers of the linear MPNN satisfies
\[
  E\bigl(H^{(\ell)}\bigr) \le (1 - \alpha \lambda_2)^{2\ell}\, E\bigl(H^{(0)}\bigr).
\]
In particular, $E(H^{(\ell)}) \to 0$ exponentially as $\ell \to \infty$: all
representations converge to a common vector.
\end{theorem}

\begin{proof}
Let $\{u_k\}_{k=1}^{n}$ be the orthonormal eigenbasis of
$L_{\mathrm{sym}} = D^{-1/2} L_G D^{-1/2}$ with eigenvalues $\lambda_k$. For each
column define $\mathbf{f}_c = D^{1/2} \mathbf{h}_c$, so that
$\mathbf{h}_c^{\!\top} L_G \mathbf{h}_c = \mathbf{f}_c^{\!\top} L_{\mathrm{sym}}
\mathbf{f}_c$. The graph Fourier transform decomposes
$\mathbf{f}_c = \sum_{k=1}^{n} \hat f_{c,k}\, u_k$.

The diffusion operator satisfies
$D^{1/2}(I - \alpha \tilde L) D^{-1/2} = I - \alpha L_{\mathrm{sym}}$, with
eigenvalues $1 - \alpha \lambda_k$. After $\ell$ iterations the spectral
coefficients of $\mathbf{f}_c^{(\ell)} = D^{1/2} \mathbf{h}_c^{(\ell)}$ scale as
$(1 - \alpha \lambda_k)^\ell \hat f_{c,k}$. Hence the Dirichlet energy in the
spectral domain is
\[
  E(H^{(\ell)})
    = \sum_{c=1}^{d} \mathbf{f}_c^{(\ell)\top} L_{\mathrm{sym}} \mathbf{f}_c^{(\ell)}
    = \sum_{c=1}^{d} \sum_{k=2}^{n} \lambda_k (1 - \alpha \lambda_k)^{2\ell}
      |\hat f_{c,k}|^2,
\]
where the $k = 1$ term vanishes because $\lambda_1 = 0$. Since
$\alpha < 1/\lambda_n$ we have $|1 - \alpha \lambda_k| < 1$ for $k \ge 2$ and in
particular $|1 - \alpha \lambda_k| \le 1 - \alpha \lambda_2$ for all $k \ge 2$
(the maximum is attained at $\lambda_2$, the lowest non-trivial frequency).
Therefore
\[
  E(H^{(\ell)})
    \le (1 - \alpha \lambda_2)^{2\ell} \sum_{c=1}^{d} \sum_{k=2}^{n}
        \lambda_k |\hat f_{c,k}|^2
    = (1 - \alpha \lambda_2)^{2\ell}\, E(H^{(0)}).
\]
Since $0 < \alpha \lambda_2 < 1$, the factor $(1 - \alpha \lambda_2)^{2\ell} \to 0$
exponentially. \qedhere
\end{proof}

\begin{remark}[Geometric interpretation of oversmoothing]
\label{rem:priors:oversmoothing}
The convergence rate $(1 - \alpha \lambda_2)^{2\ell}$ is controlled by the
\emph{spectral gap} $\lambda_2$, the second eigenvalue of the Laplacian. Graphs
with large $\lambda_2$ (such as expanders) converge faster to the uniform state,
while poorly connected graphs ($\lambda_2 \approx 0$) better preserve the
diversity of representations. This is the exact discrete counterpart of heat
diffusion on Riemannian manifolds, where the spectral gap of the Laplace--Beltrami
operator controls the mixing rate. Among the strategies to mitigate oversmoothing
are residual connections (which preserve the original signal), layer
normalization, and topological approaches such as Neural Sheaf
Diffusion~\citep{bodnar2022neural}, where scalar diffusion is replaced by a
diffusion on sheaves that lets each edge transport information in different
vector spaces.
\end{remark}

\subsection{Oversquashing: topological bottlenecks}
\label{sec:priors:oversquashing}

If oversmoothing arises from the loss of high-frequency signal, the complementary
phenomenon of \emph{oversquashing} is the inability to propagate information over
long distances. Intuitively, the information from a distant node must be
compressed exponentially to ``pass'' through the bottlenecks of the graph.

\begin{definition}[Jacobian sensitivity of an MPNN]
\label{def:priors:sensitivity}
For an MPNN with representations $h_v^{(\ell)} \in \R^d$, the \emph{sensitivity}
of node $v$ to node $u$ after $\ell$ layers is
\[
  S_{vu}^{(\ell)} = \left\| \frac{\partial h_v^{(\ell)}}{\partial h_u^{(0)}} \right\|,
\]
the operator norm of the Jacobian matrix
$\partial h_v^{(\ell)} / \partial h_u^{(0)} \in \R^{d \times d}$.
\end{definition}

If $S_{vu}^{(\ell)} \approx 0$, node $v$ is essentially insensitive to the input
of node $u$: any information $u$ carries has been lost during propagation.

\begin{proposition}[Exponential decay of sensitivity]
\label{prop:priors:oversquashing}
Let an MPNN have message and update functions that are Lipschitz with constants
$c_M$ and $c_U$ respectively, and let $\hat A = D^{-1} A$ be the random-walk
transition matrix of $\mathcal{G}$. Then for any nodes $v, u \in V$,
\[
  S_{vu}^{(\ell)} \le (c_U \cdot c_M)^\ell\, [\hat A^\ell]_{vu}.
\]
In particular, if the graph distance satisfies $d_{\mathcal{G}}(v, u) > \ell$,
then $[\hat A^\ell]_{vu} = 0$ and the sensitivity is exactly zero.
\end{proposition}

\begin{proof}[Proof sketch]
We proceed by induction on $\ell$. For $\ell = 1$ and nodes $v \neq u$, since
$h_v^{(1)} = U(h_v^{(0)}, m_v^{(1)})$, the chain rule gives
\[
  \frac{\partial h_v^{(1)}}{\partial h_u^{(0)}}
    = \frac{\partial U}{\partial h_v^{(0)}}
      \cdot \underbrace{\frac{\partial h_v^{(0)}}{\partial h_u^{(0)}}}_{=\,0 \text{ if } v \neq u}
    + \frac{\partial U}{\partial m_v} \cdot \frac{\partial m_v}{\partial h_u^{(0)}}
    = \frac{\partial U}{\partial m_v} \cdot \frac{\partial m_v}{\partial h_u^{(0)}}.
\]
The second factor is nonzero only if $u \in \Nb(v)$, in which case the Lipschitz
constant of $M$ contributes $c_M$ and the degree normalization contributes
$1/\deg(v) = [\hat A]_{vu}$. Thus $S_{vu}^{(1)} \le c_U c_M [\hat A]_{vu}$.

For the inductive step, the multivariable chain rule gives
\[
  \frac{\partial h_v^{(\ell+1)}}{\partial h_u^{(0)}}
    = \sum_{w \in \Nb(v) \cup \{v\}}
      \frac{\partial h_v^{(\ell+1)}}{\partial h_w^{(\ell)}}
      \cdot \frac{\partial h_w^{(\ell)}}{\partial h_u^{(0)}}.
\]
Bounding each factor by the Lipschitz constants and using the inductive
hypothesis for $\partial h_w^{(\ell)} / \partial h_u^{(0)}$, we obtain the matrix
product $\hat A^{\ell+1}$ that records all paths of length $\ell + 1$ from $u$ to
$v$. \qedhere
\end{proof}

\begin{remark}[Curvature and oversquashing]
\label{rem:priors:curvature}
The bound reveals that the topology of the graph, encoded in $\hat A^\ell$, is
the determining factor of oversquashing. \citet{topping2022understanding}
formalize this connection through the \emph{discrete Ricci curvature} (Ollivier):
edges $(u, v)$ with negative Ricci curvature correspond to bottlenecks where
$[\hat A^\ell]_{vu}$ decays rapidly. This is the discrete counterpart of the
Jacobi equation in Riemannian geometry: negative sectional curvature makes
geodesics diverge, reducing the ``volume of information'' that can propagate.
Topping et al.\ propose \emph{rewiring} the graph via Ricci flow, adding edges in
regions of negative curvature to relieve the bottlenecks --- an example of how
differential geometry prescribes solutions to GNN problems.
\end{remark}

\subsection{The k-WL hierarchy and extensions}
\label{sec:priors:kwl}

\Cref{thm:priors:wl-mpnn} establishes that MPNNs are as expressive as the 1-WL
test, which operates on individual nodes. A natural extension considers
higher-order tests operating on tuples of nodes.

\begin{definition}[Weisfeiler--Leman test of order $k$]
\label{def:priors:kwl}
The \emph{$k$-WL test} operates on $k$-tuples of nodes
$\bar v = (v_1, \dots, v_k) \in V^k$~\citep{morris2019weisfeiler}. One defines an
initial colouring $c^{(0)}(\bar v)$ based on the isomorphism type of the
subtuple, and the iterative refinement
\[
  c^{(t+1)}(\bar v) = \HASH\Bigl(c^{(t)}(\bar v),\,
    \bigl\{\!\!\bigl\{ \bigl(c^{(t)}(\bar v[w/1]), \dots,
      c^{(t)}(\bar v[w/k])\bigr) : w \in V \bigr\}\!\!\bigr\}\Bigr),
\]
where $\bar v[w/i]$ denotes the tuple obtained by replacing the $i$-th component
of $\bar v$ by $w$.
\end{definition}

\begin{proposition}[Strict expressivity hierarchy]
\label{prop:priors:kwl-hierarchy}
The Weisfeiler--Leman tests form a strictly increasing hierarchy in expressive
power,
\[
  1\text{-WL} \subsetneq 2\text{-WL} \subsetneq 3\text{-WL} \subsetneq \cdots,
\]
that is, for every $k \ge 1$ there exist graphs that $k$-WL does not distinguish
but $(k+1)$-WL does. The computational complexity of the $k$-WL test is
$O(n^k \log n)$ per iteration, where $n = |V|$.
\end{proposition}

\begin{proof}
The proof of strict separation requires techniques from finite model theory (in
particular the connection with the $C^k$ logic of Immerman--Lander) and exceeds
the scope of this book; see \citet{morris2019weisfeiler} for details. The
complexity $O(n^k \log n)$ follows directly from the iterative refinement: there
are $n^k$ tuples, each iteration inspects $n$ substitutions per tuple, and
stabilization occurs in at most $O(n^k)$ iterations with $O(\log n)$ hashing
cost. \qedhere
\end{proof}

Several strategies overcome the 1-WL barrier without incurring the exponential
cost of $k$-WL:
\begin{itemize}
\item \textbf{Simplicial message passing}: as
\Cref{prop:priors:simplicial-expressivity} shows, operating on simplicial
complexes ($k$-chains) breaks 1-WL by incorporating higher-order topological
information.
\item \textbf{Structural encodings}: adding positional encodings based on
Laplacian eigenvectors or random walks breaks the symmetry of the nodes.
\item \textbf{Spectral graph transformers}: architectures such as that of
\citet{kreuzer2021spectral} combine global attention with spectral encodings,
connecting spectral analysis with attention mechanisms.
\end{itemize}

The three limitations analysed (oversmoothing, oversquashing, and the WL barrier)
are manifestations of the fundamentally \emph{local} character of message
passing: each layer propagates information only one hop. Oversmoothing shows that
too many layers destroy the discriminative signal; oversquashing shows that too
few prevent long-range communication; and the WL barrier shows that certain
global patterns are undetectable locally. The geometric tools developed in this
book (spectral analysis, discrete curvature, higher-order topology) do not merely
\emph{diagnose} these limitations: they \emph{prescribe} concrete solutions.

\section{Summary}
\label{sec:priors:summary}

Invariance and equivariance turn the informal demand that a model ``respect the
symmetries of the data'' into precise constraints on a function. Because
equivariant maps compose (\Cref{thm:priors:composition}), these constraints yield
a constructive blueprint: stack equivariant layers and local coarsening, finish
with an invariant readout. The equivariant universal approximation theorem
(\Cref{thm:priors:universal}) certifies that this discipline costs no expressive
power. Message passing makes the blueprint concrete and reveals fully connected,
convolutional, attention-based, graph, and even simplicial architectures as one
construction under different geometric priors, with the WL hierarchy,
oversmoothing, and oversquashing delimiting its expressive reach. The unifying
perspective of GDL exposes a natural progression in geometric sophistication:
fully connected networks operate in Euclidean spaces with trivial symmetries;
convolutions introduce a spatial inductive bias through locality and translation
equivariance; transformers generalise this through adaptive global connectivity;
and the paradigm extends naturally to arbitrary graphs and higher-order
topologies. This framework is predictive, not merely descriptive: it suggests
design principles for new hybrid architectures. The remaining chapters
instantiate this blueprint on each of the five geometric domains.
