\providecommand{\SO}{\mathrm{SO}}
\providecommand{\SE}{\mathrm{SE}}
\providecommand{\Eucl}{\mathrm{E}}
\providecommand{\Hyp}{\mathbb{H}}
\providecommand{\Attn}{\operatorname{Attention}}
\providecommand{\softmax}{\operatorname{softmax}}
\providecommand{\Concat}{\operatorname{Concat}}
\providecommand{\sigm}{\sigma}
\providecommand{\Lc}{\mathcal{L}}

\chapter{Applications}
\label{ch:applications}

The previous chapters built a theory: domains carry symmetries, symmetries
constrain architectures, and a single principle --- equivariant message
passing --- unifies convolutional networks, graph networks, and attention
mechanisms. This chapter turns to practice. Its purpose is not to catalogue
every system that uses geometric deep learning, but to show how the abstract
machinery of the ``5~Gs'' becomes concrete in real problems, and why the
geometric viewpoint translates into measurable gains in accuracy, data
efficiency, and physical consistency.

The chapter has two movements. The first develops the \emph{attention
mechanism} in full: its history, its algebraic foundations, the geometry of its
query, key, and value spaces, its formulation as an integral operator, and ---
crucially --- the precise sense in which it is the adaptive counterpart of the
geometric convolution of \Cref{ch:grids,ch:groups}. This material is foundational
rather than merely applied: the transformer is the architecture behind much of
modern machine learning, and understanding it geometrically explains both its
success and the long line of equivariant variants that power the rest of the
chapter. The second movement surveys the application areas themselves,
organized around three families of data. \emph{Computer vision} ranges from
regular pixel grids to unstructured three-dimensional point clouds, exercising
the grid, group, and gauge themes. \emph{Molecules and physical systems} are the
showcase of $\Eucl(3)$ and $\SE(3)$ equivariance, where respecting the
symmetries of physics is not a refinement but a requirement. \emph{Recommender
systems and social networks} are relational by nature and rely on permutation
invariance and, increasingly, on non-Euclidean geometry.

\section{Attention mechanisms}
\label{sec:apps:attention-foundations}

Attention mechanisms are the natural complement of the geometric convolutions
developed in the previous chapters: where convolutions impose rigid structure
through shared, content-independent kernels, attention introduces adaptive
flexibility, weighting the contribution of each element dynamically according to
the information it carries. This section develops the mathematical foundations
of attention, from its classical algebraic formulation to the geometric
extensions that let it operate on non-Euclidean domains while preserving
symmetries.

\subsection{History and motivation}
\label{subsec:apps:attn-history}

Attention mechanisms emerged historically from the need to overcome the
limitations of sequential models in natural language processing. The early work
of \citet{bahdanau2014neural} and \citet{luong2015effective} introduced
attention into recurrent-network machine-translation systems, allowing the model
to ``attend'' to different parts of the input sequence during generation. The
turning point came with the seminal \emph{``Attention is All You Need''} of
\citet{vaswani2017attentionneed}, which showed that attention alone --- without
recurrence or convolution --- could reach state-of-the-art performance on
translation. This transformer architecture revolutionized not only natural
language processing but became the foundation of models for computer vision,
graph processing, and, eventually, geometric deep learning.

The historical evolution can be summarized in three phases: attention as a
\emph{complement} to recurrent networks (2014--2016), used to improve the
handling of long-range dependencies; attention as the \emph{principal
architecture} (2017--2019), with pure transformers replacing recurrent models;
and \emph{geometric attention} (2020--present), extending the mechanism to
non-Euclidean data while preserving symmetries. The remainder of this chapter
follows that arc to its current destination.

\subsection{Algebraic foundations}
\label{subsec:apps:attn-algebra}

Attention rests on concepts from linear algebra and the geometry of vector
spaces. In essence it operates on three interrelated vector spaces that encode
different aspects of the information.

\paragraph{Feature spaces.}
Consider a feature space $\mathcal{H} = \R^d$ in which the input
representations live. For a set of $n$ input elements we collect them into a
matrix
\[
  X = [x_1, x_2, \dots, x_n]^\top \in \R^{n \times d},
\]
where each $x_i \in \mathcal{H}$ is one element of the input sequence. The
attention mechanism defines three fundamental linear transformations, the
\emph{queries}, \emph{keys}, and \emph{values},
\[
  Q = X W_Q \in \R^{n\times d_k},
  \qquad
  K = X W_K \in \R^{n\times d_k},
  \qquad
  V = X W_V \in \R^{n\times d_v},
\]
where the weight matrices $W_Q, W_K \in \R^{d\times d_k}$ and
$W_V \in \R^{d\times d_v}$ are trainable parameters projecting the input space
onto specialized subspaces. In the standard multi-head implementation one takes
$d_k = d_v = d/h$, where $h$ is the number of attention heads.

\begin{definition}[Dot product as a similarity measure]
\label{def:apps:dot-similarity}
The fundamental compatibility function in attention is the dot product
\[
  \operatorname{sim}(q_i, k_j) = q_i^\top k_j
  = \norm{q_i}\,\norm{k_j}\cos\theta_{ij},
\]
where $\theta_{ij}$ is the angle between $q_i$ and $k_j$.
\end{definition}

This geometric reading shows that attention measures a normalized cosine
similarity between representations in feature space: aligned query--key pairs
score highly, orthogonal pairs score near zero.

\paragraph{The geometry of the Q, K, V spaces.}
The compatibility of all query--key pairs is collected in the
\emph{similarity matrix}
\[
  S = Q K^\top, \qquad S_{ij} = \inner{q_i}{k_j} \in \R^{n\times n}.
\]
This matrix encodes all the geometric relations between queries and keys; when
$Q = K$ (pure self-attention) it coincides with the Gram matrix of the keys. It
admits a singular value decomposition $S = U\Sigma V^\top$ whose factors capture
the principal directions of queries and keys, its Frobenius norm satisfies
$\norm{S}_F^2 = \operatorname{tr}(QK^\top K Q^\top)$, and entrywise
$S_{ij} = \norm{q_i}\norm{k_j}\cos\theta_{ij}$.

\begin{theorem}[Geometric projection in attention]
\label{thm:apps:geom-projection}
The attention mechanism can be interpreted as a geometric projection of the
value space onto the subspace spanned by the queries, mediated by the metric
induced by the keys. For each query $q_i$,
\[
  y_i = \sum_{j=1}^{n} \alpha_{ij}\, v_j = V^\top \boldsymbol{\alpha}_i,
  \qquad
  \boldsymbol{\alpha}_i = \softmax\!\bigl(S_{i,:}/\sqrt{d_k}\bigr),
\]
where $\boldsymbol{\alpha}_i$ is a probability distribution over the value
index.
\end{theorem}

\begin{proof}
The attention weights $\alpha_{ij}$ form a convex (probabilistic) combination
over the values, determined by the geometric proximity of $q_i$ to each key
$k_j$ as measured by the scaled, softmax-normalized inner product. The resulting
projection preserves the structure of the value space while emphasizing
directions aligned with the query according to the metric defined by the keys.
\end{proof}

\Cref{fig:apps:qkv} illustrates this two-stage picture: similarities are read
off as angles in the query--key space, and the output is the corresponding
convex combination of value vectors.

\begin{figure}[h]
\centering
\begin{tikzpicture}[>=Stealth, font=\small, scale=0.82, every node/.style={transform shape}]
% ============================================================
% LEFT PANEL: Query-Key space (dot-product similarity)
% ============================================================
\begin{scope}[shift={(0,0)}]
  \node[font=\bfseries, align=center, text=black] at (3.2, 6.8)
    {Query and key space};
  \draw[->, gdlgray!60, line width=0.8pt] (-0.3, 0) -- (6.8, 0)
    node[right, font=\footnotesize, text=gdlgray!70!black] {$e_1$};
  \draw[->, gdlgray!60, line width=0.8pt] (0, -0.3) -- (0, 6.5)
    node[above, font=\footnotesize, text=gdlgray!70!black] {$e_2$};
  \foreach \x in {1,...,6} {\draw[gdlgray!15] (\x, -0.1) -- (\x, 6.3);}
  \foreach \y in {1,...,6} {\draw[gdlgray!15] (-0.1, \y) -- (6.6, \y);}
  \node[font=\footnotesize, text=gdlgray!70!black, anchor=north east] at (-0.1, -0.1) {$0$};
  \draw[gdlpurple!60, line width=0.8pt]
    ({2.2*cos(42.0)}, {2.2*sin(42.0)}) arc[start angle=42.0, end angle=51.1, radius=2.2];
  \node[font=\scriptsize, text=gdlpurple!70!black, anchor=south west]
    at ({2.55*cos(46.5)}, {2.55*sin(46.5)}) {$\theta_{i1}$};
  \draw[gdlpurple!60, line width=0.8pt]
    ({1.8*cos(23.3)}, {1.8*sin(23.3)}) arc[start angle=23.3, end angle=42.0, radius=1.8];
  \node[font=\scriptsize, text=gdlpurple!70!black, anchor=north west]
    at ({2.15*cos(32.5)}, {2.15*sin(32.5)}) {$\theta_{i2}$};
  \draw[gdlpurple!60, line width=0.8pt]
    ({1.5*cos(42.0)}, {1.5*sin(42.0)}) arc[start angle=42.0, end angle=78.3, radius=1.5];
  \node[font=\scriptsize, text=gdlpurple!70!black, anchor=south]
    at ({1.9*cos(60)}, {1.9*sin(60)}) {$\theta_{i3}$};
  \draw[-{Stealth[length=5pt,width=4pt]}, gdlblue!80, line width=1.8pt] (0,0) -- (5.0, 4.5);
  \node[font=\small, text=gdlblue!70!black, anchor=south west] at (5.1, 4.5) {$\mathbf{q}_i$};
  \fill[gdlblue!70] (5.0, 4.5) circle (3pt);
  \draw[-{Stealth[length=5pt,width=4pt]}, gdlred!80, line width=1.8pt] (0,0) -- (4.2, 5.2);
  \node[font=\small, text=gdlred!70!black, anchor=south] at (4.2, 5.35) {$\mathbf{k}_1$};
  \fill[gdlred!70] (4.2, 5.2) circle (3pt);
  \draw[-{Stealth[length=5pt,width=4pt]}, gdlred!80, line width=1.8pt] (0,0) -- (5.8, 2.5);
  \node[font=\small, text=gdlred!70!black, anchor=west] at (5.9, 2.5) {$\mathbf{k}_2$};
  \fill[gdlred!70] (5.8, 2.5) circle (3pt);
  \draw[-{Stealth[length=5pt,width=4pt]}, gdlred!80, line width=1.8pt] (0,0) -- (1.2, 5.8);
  \node[font=\small, text=gdlred!70!black, anchor=south] at (1.2, 5.9) {$\mathbf{k}_3$};
  \fill[gdlred!70] (1.2, 5.8) circle (3pt);
  \coordinate (projfoot) at (4.17, 5.17);
  \draw[gdlpurple!40, line width=0.6pt, densely dotted] (5.0, 4.5) -- (projfoot);
  \fill[gdlpurple!50] (projfoot) circle (2pt);
  \node[font=\tiny, text=gdlpurple!70!black, fill=white, fill opacity=0.85,
        text opacity=1, inner sep=1.5pt, rounded corners=1pt, anchor=south west]
    at (4.5, 5.35) {high similarity};
  \node[font=\tiny, text=gdlpurple!70!black, fill=white, fill opacity=0.85,
        text opacity=1, inner sep=1.5pt, rounded corners=1pt, anchor=south west]
    at (6.0, 2.55) {medium};
  \node[font=\tiny, text=gdlpurple!70!black, fill=white, fill opacity=0.85,
        text opacity=1, inner sep=1.5pt, rounded corners=1pt, anchor=south east]
    at (0.9, 5.95) {low};
  \node[font=\scriptsize, text=gdlpurple!70!black, align=center, anchor=north west]
    at (0.3, -0.8) {$\alpha_{ij} = \operatorname{softmax}_j\!\left(\dfrac{\mathbf{q}_i \cdot \mathbf{k}_j}{\sqrt{d_k}}\right)$};
\end{scope}
% ============================================================
% ARROW between panels
% ============================================================
\draw[->, gdlorange!70, line width=2pt] (7.8, 3.2) -- (9.5, 3.2);
\node[font=\footnotesize, align=center, text=gdlorange!70!black] at (8.65, 4.0)
  {Attention\\weights $\alpha_{ij}$};
% ============================================================
% RIGHT PANEL: Weighted value aggregation
% ============================================================
\begin{scope}[shift={(10.0, 0)}]
  \node[font=\bfseries, align=center, text=black] at (3.5, 6.8)
    {Weighted aggregation of values};
  \draw[->, gdlgray!60, line width=0.8pt] (-0.3, 0) -- (7.3, 0)
    node[right, font=\footnotesize, text=gdlgray!70!black] {$e_1$};
  \draw[->, gdlgray!60, line width=0.8pt] (0, -0.3) -- (0, 6.5)
    node[above, font=\footnotesize, text=gdlgray!70!black] {$e_2$};
  \foreach \x in {1,...,7} {\draw[gdlgray!15] (\x, -0.1) -- (\x, 6.3);}
  \foreach \y in {1,...,6} {\draw[gdlgray!15] (-0.1, \y) -- (7.1, \y);}
  \node[font=\footnotesize, text=gdlgray!70!black, anchor=north east] at (-0.1, -0.1) {$0$};
  \draw[-{Stealth[length=4pt,width=3pt]}, gdlgreen!35, line width=0.8pt] (0,0) -- (3.0, 5.5);
  \node[font=\small, text=gdlgreen!70!black, anchor=south] at (3.0, 5.6) {$\mathbf{v}_1$};
  \fill[gdlgreen!60] (3.0, 5.5) circle (2.5pt);
  \draw[-{Stealth[length=4pt,width=3pt]}, gdlgreen!35, line width=0.8pt] (0,0) -- (5.5, 2.0);
  \node[font=\small, text=gdlgreen!70!black, anchor=north west] at (5.6, 1.9) {$\mathbf{v}_2$};
  \fill[gdlgreen!60] (5.5, 2.0) circle (2.5pt);
  \draw[-{Stealth[length=3pt,width=2.5pt]}, gdlgreen!25, line width=0.6pt] (0,0) -- (1.5, 4.0);
  \node[font=\small, text=gdlgreen!70!black, anchor=east] at (1.3, 4.3) {$\mathbf{v}_3$};
  \fill[gdlgreen!50] (1.5, 4.0) circle (2pt);
  \draw[-{Stealth[length=5pt,width=4pt]}, gdlgreen!75, line width=2.2pt] (0,0) -- (1.95, 3.575);
  \node[font=\scriptsize, text=gdlgreen!70!black, anchor=east,
        fill=white, fill opacity=0.85, text opacity=1, inner sep=1pt, rounded corners=1pt]
    at (1.5, 2.5) {$\alpha_{i1}\mathbf{v}_1$};
  \draw[-{Stealth[length=4pt,width=3pt]}, gdlgreen!65, line width=1.4pt] (0,0) -- (1.375, 0.5);
  \node[font=\scriptsize, text=gdlgreen!70!black, anchor=north west,
        fill=white, fill opacity=0.85, text opacity=1, inner sep=1pt, rounded corners=1pt]
    at (1.45, 0.55) {$\alpha_{i2}\mathbf{v}_2$};
  \draw[-{Stealth[length=3pt,width=2pt]}, gdlgreen!50, line width=0.6pt] (0,0) -- (0.15, 0.4);
  \draw[gdlorange!60, line width=0.7pt, dashed] (1.95, 3.575) -- (3.325, 4.075);
  \draw[gdlorange!60, line width=0.7pt, dashed] (1.375, 0.5) -- (3.325, 4.075);
  \draw[gdlorange!50, line width=0.5pt, densely dotted] (3.325, 4.075) -- (3.475, 4.475);
  \draw[-{Stealth[length=6pt,width=5pt]}, gdlorange!80, line width=2.4pt] (0,0) -- (3.475, 4.475);
  \node[font=\small\bfseries, text=gdlorange!70!black, anchor=south west] at (3.55, 4.5) {$\mathbf{y}_i$};
  \fill[gdlorange!80] (3.475, 4.475) circle (3.5pt);
  \node[font=\scriptsize, text=gdlpurple!70!black, fill=white, fill opacity=0.85,
        text opacity=1, inner sep=2pt, rounded corners=1pt, anchor=west]
    at (3.3, 5.5) {$\alpha_{i1} = 0.65$};
  \node[font=\scriptsize, text=gdlpurple!70!black, fill=white, fill opacity=0.85,
        text opacity=1, inner sep=2pt, rounded corners=1pt, anchor=west]
    at (5.8, 2.5) {$\alpha_{i2} = 0.25$};
  \node[font=\scriptsize, text=gdlpurple!70!black, fill=white, fill opacity=0.85,
        text opacity=1, inner sep=2pt, rounded corners=1pt, anchor=east]
    at (1.0, 4.6) {$\alpha_{i3} = 0.10$};
  \node[font=\scriptsize, text=gdlorange!70!black, align=center, anchor=north west]
    at (0.3, -0.8) {$\mathbf{y}_i = \displaystyle\sum_{j} \alpha_{ij}\, \mathbf{v}_j$};
\end{scope}
\end{tikzpicture}
\caption[Geometric interpretation of attention]{Geometric interpretation of the
attention mechanism. \textbf{Left}: in the query and key space, the similarity
between $\mathbf{q}_i$ and each $\mathbf{k}_j$ is measured by the angle
$\theta_{ij}$ (scaled inner product). \textbf{Right}: the values
$\mathbf{v}_j$ are weighted by the attention coefficients $\alpha_{ij}$, and the
output $\mathbf{y}_i$ is their convex combination.}
\label{fig:apps:qkv}
\end{figure}

\subsection{The basic attention mechanism}
\label{subsec:apps:attn-basic}

Scaled dot-product attention is defined formally as
\[
  \Attn(Q, K, V) = \softmax\!\left(\frac{QK^\top}{\sqrt{d_k}}\right) V .
\]
The score matrix $S = QK^\top \in \R^{n\times n}$ encodes the compatibility of
each query--key pair, and the scale factor $1/\sqrt{d_k}$ prevents the dot
products from growing excessively with the dimension and saturating the softmax.
Equivalently, for each element
\[
  y_i = \sum_{j=1}^{n} \alpha_{ij}\, v_j,
  \qquad \sum_{j=1}^{n}\alpha_{ij} = 1, \qquad \alpha_{ij}\ge 0,
\]
so the output is a convex interpolation in the value space, with weights forming
a probability distribution over the inputs.

\begin{definition}[Generalized attention function]
\label{def:apps:generalized-attn}
A \emph{generalized attention function} computes, for each query,
\[
  y_i = \sum_{j=1}^{n}\alpha_{ij}\, v_j,
  \qquad
  \alpha_{ij} = \frac{\exp(e_{ij})}{\sum_{k=1}^n \exp(e_{ik})},
  \qquad
  e_{ij} = a(q_i, k_j),
\]
where $a:\mathcal{X}\times\mathcal{Y}\to\R$ is a compatibility function. The
scaled dot product is the choice $a(q,k) = \inner{q}{k}/\sqrt{d_k}$, but any
$G$-invariant scalar of the pair $(q_i, k_j)$ is admissible --- the freedom that
the geometric variants below exploit.
\end{definition}

\emph{Self-attention} is the special case $Q = K = V = X$ in which queries,
keys, and values all come from the same input, letting every element attend to
all the others and capturing global dependencies in a single operation. This
construction is also maximally expressive among permutation-equivariant maps.

\begin{theorem}[Universality of self-attention]
\label{thm:apps:self-attn-universal}
Let $\mathcal{K}\subseteq\R^{n\times d}$ be compact. Any continuous
permutation-equivariant function $f:\mathcal{K}\to\R^{n\times d}$ (that is,
$f(PX) = P f(X)$ for every permutation matrix $P$) can be approximated uniformly
on $\mathcal{K}$ to arbitrary precision. Full universality requires the complete
transformer block --- self-attention together with feed-forward layers and
residual connections --- not isolated self-attention layers alone; see
\citet{yun2020transformers} for the rigorous statement.
\end{theorem}

\begin{proof}[Sketch]
By the representation theorem for symmetric functions~\citep{zaheer2017deepsets},
any continuous permutation-equivariant $f$ can be written as
$f(X)_i = \phi\bigl(x_i, \sum_j \rho(x_j)\bigr)$ for suitable continuous
$\phi,\rho$. By the universal approximation theorem one approximates this, for
any $\epsilon>0$, by a sum of pairwise interactions
$\sum_j \softmax\bigl(g_1(x_i)^\top g_2(x_j)\bigr)\, h(g_3(x_j))$, which is
realized exactly by a self-attention layer whose projection matrices implement
$g_1,g_2,g_3$, the scale factor being absorbed into $g_1,g_2$. Composition of
universally approximable maps gives the result.
\end{proof}

\subsection{Multi-head attention}
\label{subsec:apps:multihead}

Multi-head attention extends the basic mechanism by processing several
representation subspaces in parallel,
\[
  \operatorname{MultiHead}(Q,K,V) = \Concat(\mathrm{head}_1,\dots,\mathrm{head}_h)\,W_O,
  \qquad
  \mathrm{head}_i = \Attn(QW_Q^i, KW_K^i, VW_V^i),
\]
with projections $W_Q^i, W_K^i \in \R^{d\times d_k}$ and
$W_V^i \in \R^{d\times d_v}$ for $i = 1,\dots,h$ and an output projection
$W_O \in \R^{hd_v\times d}$. The decomposition into parallel subspaces lets each
head specialize in a different type of relation, increases the model's capacity
to attend to several positions and subspaces at once, and improves training
stability through partial redundancy (\Cref{fig:apps:multihead}).

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    scale=0.78, every node/.style={transform shape},
    arrow/.style={-{Stealth[length=5pt]}, thick, draw=gdlgray!60},
    line/.style={thick, draw=gdlgray!60},
    inputbox/.style={rectangle, rounded corners=3pt, draw=gdlblue!60, fill=gdlblue!20,
        thick, minimum width=2.4cm, minimum height=0.7cm, font=\small,
        text=gdlblue!70!black, align=center},
    projbox/.style={rectangle, rounded corners=3pt, draw=gdlgreen!60, fill=gdlgreen!20,
        thick, minimum width=1.1cm, minimum height=0.55cm, font=\footnotesize,
        text=gdlgreen!70!black, align=center},
    attnbox/.style={rectangle, rounded corners=3pt, draw=gdlorange!60, fill=gdlorange!20,
        thick, minimum width=2.8cm, minimum height=0.6cm, font=\footnotesize,
        text=gdlorange!70!black, align=center},
    concatbox/.style={rectangle, rounded corners=3pt, draw=gdlpurple!60, fill=gdlpurple!20,
        thick, minimum width=3.2cm, minimum height=0.7cm, font=\small,
        text=gdlpurple!70!black, align=center},
    outputbox/.style={rectangle, rounded corners=3pt, draw=gdlblue!60, fill=gdlblue!20,
        thick, minimum width=2.4cm, minimum height=0.7cm, font=\small,
        text=gdlblue!70!black, align=center},
    headlabel/.style={font=\footnotesize\bfseries, text=gdlgray!70!black},
    dashedbox/.style={rectangle, rounded corners=5pt, draw=gdlgray!40, dashed,
        thick, inner sep=10pt, inner ysep=10pt}
]
\def\headspacing{4.0}
\def\projy{-2.8}
\def\attny{-4.2}
\def\distribY{-1.2}
\def\concaty{-6.0}
\def\outputy{-7.5}
\node[inputbox] (input) at (0, 0) {$\mathbf{X} \in \mathbb{R}^{n \times d}$};
\node[projbox] (q1) at ({-\headspacing - 0.65}, \projy) {$\mathbf{Q}_1$};
\node[projbox] (k1) at ({-\headspacing + 0.35}, \projy) {$\mathbf{K}_1$};
\node[projbox] (v1) at ({-\headspacing + 1.35}, \projy) {$\mathbf{V}_1$};
\node[attnbox] (attn1) at ({-\headspacing + 0.35}, \attny)
    {$\mathrm{Attn}(\mathbf{Q}_1, \mathbf{K}_1, \mathbf{V}_1)$};
\node[projbox] (q2) at ({-0.65}, \projy) {$\mathbf{Q}_2$};
\node[projbox] (k2) at (0.35, \projy) {$\mathbf{K}_2$};
\node[projbox] (v2) at (1.35, \projy) {$\mathbf{V}_2$};
\node[attnbox] (attn2) at (0.35, \attny)
    {$\mathrm{Attn}(\mathbf{Q}_2, \mathbf{K}_2, \mathbf{V}_2)$};
\node[projbox] (q3) at ({\headspacing - 0.65}, \projy) {$\mathbf{Q}_3$};
\node[projbox] (k3) at ({\headspacing + 0.35}, \projy) {$\mathbf{K}_3$};
\node[projbox] (v3) at ({\headspacing + 1.35}, \projy) {$\mathbf{V}_3$};
\node[attnbox] (attn3) at ({\headspacing + 0.35}, \attny)
    {$\mathrm{Attn}(\mathbf{Q}_3, \mathbf{K}_3, \mathbf{V}_3)$};
\foreach \h in {1,2,3} {
    \draw[arrow] (q\h.south) -- (attn\h.north -| q\h.south);
    \draw[arrow] (k\h.south) -- (attn\h.north -| k\h.south);
    \draw[arrow] (v\h.south) -- (attn\h.north -| v\h.south);
}
\node[dashedbox, fit=(q1)(v1)(attn1)] (hbox1) {};
\node[dashedbox, fit=(q2)(v2)(attn2)] (hbox2) {};
\node[dashedbox, fit=(q3)(v3)(attn3)] (hbox3) {};
\node[headlabel, anchor=south] at (hbox1.north) {Head 1};
\node[headlabel, anchor=south] at (hbox2.north) {Head 2};
\node[headlabel, anchor=south] at (hbox3.north) {Head 3};
\draw[line] (input.south) -- (0, \distribY);
\draw[line] ({-\headspacing + 0.35}, \distribY) -- ({\headspacing + 0.35}, \distribY);
\pgfmathsetmacro{\arrowEndY}{\projy + 1.45}
\draw[arrow] ({-\headspacing + 0.35}, \distribY) -- ({-\headspacing + 0.35}, \arrowEndY);
\draw[arrow] (0.35, \distribY) -- (0.35, \arrowEndY);
\draw[arrow] ({\headspacing + 0.35}, \distribY) -- ({\headspacing + 0.35}, \arrowEndY);
\node[concatbox] (concat) at (0.35, \concaty)
    {$\mathrm{Concat}(\mathrm{head}_1, \ldots, \mathrm{head}_h)$};
\pgfmathsetmacro{\convY}{\concaty + 0.7}
\draw[line] (attn1.south) -- (attn1.south |- 0, \convY);
\draw[arrow] (attn1.south |- 0, \convY) -| ($(concat.north) + (-0.5, 0)$);
\draw[arrow] (attn2.south) -- (concat.north);
\draw[line] (attn3.south) -- (attn3.south |- 0, \convY);
\draw[arrow] (attn3.south |- 0, \convY) -| ($(concat.north) + (0.5, 0)$);
\node[outputbox] (output) at (0.35, \outputy) {$\mathbf{W}^O \cdot \mathrm{Concat}$};
\draw[arrow] (concat.south) -- (output.north);
\node[font=\small, text=gdlblue!70!black, below=4pt of output]
    {Output $\in \mathbb{R}^{n \times d}$};
\node[font=\scriptsize, text=gdlgray!70!black, align=left, anchor=north west]
    at ($(hbox3.east) + (0.5, 0.4)$)
    {$\mathbf{W}_i^Q, \mathbf{W}_i^K \in \mathbb{R}^{d \times d_k}$\\[2pt]
     $\mathbf{W}_i^V \in \mathbb{R}^{d \times d_v}$\\[2pt]
     $\mathbf{W}^O \in \mathbb{R}^{hd_v \times d}$\\[2pt]
     $d_k = d_v = d/h$};
\end{tikzpicture}
\caption[Multi-head attention]{Multi-head attention. The input is projected into
$h$ parallel subspaces, each with its own matrices $\mathbf{W}^Q_i$,
$\mathbf{W}^K_i$, $\mathbf{W}^V_i$. Each head computes attention independently,
and the results are concatenated and projected through $\mathbf{W}^O$.}
\label{fig:apps:multihead}
\end{figure}

\begin{proposition}[Representational capacity of multi-head pre-softmax scores]
\label{prop:apps:multihead-capacity}
Multi-head attention with $h$ heads can represent any pre-softmax score pattern
that decomposes as a sum of $h$ low-rank terms. Formally, if
$S^\star = \sum_{i=1}^h U_i\Sigma_i V_i^\top$ with each
$U_i\Sigma_i V_i^\top$ of rank $r_i\le d_k$, then there is a configuration with
$h$ heads whose pre-softmax score matrix is exactly $S^\star$.
\end{proposition}

\begin{proof}
Set $W_Q^i = U_i\Sigma_i^{1/2}$, $W_K^i = V_i\Sigma_i^{1/2}$, and $W_V^i = I$
for each term. Then head $i$ has pre-softmax scores
$Q_i K_i^\top = X U_i\Sigma_i V_i^\top X^\top$, and the sum over heads reproduces
$S^\star$.
\end{proof}

\begin{remark}
\label{rem:apps:multihead-nonlinearity}
This construction operates at the level of pre-softmax scores. In practice each
head applies softmax before aggregating with the values, which forces the
effective attention weights to be non-negative and to sum to one per row. This
nonlinearity prevents the multi-head mechanism from representing arbitrary
attention functions exactly at the level of the output, but the capacity result
explains why a handful of heads already captures rich, multi-relational
structure. The time and space cost of the layer is
$O(n^2 d + n d h d_k)$ and $O(n^2 + ndh)$ respectively; the $n^2$ term dominates
for long sequences and motivates sparse and linear approximations.
\end{remark}

\subsection{Attention as an integral operator}
\label{subsec:apps:integral-operator}

Self-attention can be formalized rigorously as an integral operator acting on
functions over a manifold, a viewpoint that connects it to the classical theory
of compact operators and exposes its spectral behaviour. Let $\mathcal{M}$ be a
compact Riemannian manifold equipped with a measure $\mu$. For a function
$f:\mathcal{M}\to\R^d$ the attention operator is
\[
  \mathcal{T}f(x) = \int_{\mathcal{M}} \kappa(x,y)\, f(y)\, d\mu(y),
  \qquad
  \kappa(x,y) = \sigma\!\left(\frac{\inner{Q(x)}{K(y)}}{\sqrt{d_k}}\right) V(y)^\top,
\]
with $Q,K,V$ mapping $\mathcal{M}$ into the query, key, and value spaces and
$\sigma$ a generalized softmax. The discrete transformer is the empirical
quadrature of this integral over $n$ sample points.

\begin{proposition}[Compactness of the attention operator]
\label{prop:apps:attn-compact}
If the kernel $\kappa(x,y)$ is continuous on $\mathcal{M}\times\mathcal{M}$ and
$\mathcal{M}$ is compact, then $\mathcal{T}: L^2(\mathcal{M})\to L^2(\mathcal{M})$
is compact.
\end{proposition}

\begin{proof}[Sketch]
For $f$ in the unit ball of $L^2(\mathcal{M})$, the image $\mathcal{T}f$ is
uniformly bounded (since $\kappa$ attains its maximum on the compact
$\mathcal{M}\times\mathcal{M}$) and equicontinuous (since $\kappa$ is uniformly
continuous there). By Arzel\`a--Ascoli the image of the unit ball is relatively
compact in $C(\mathcal{M})$, hence in $L^2(\mathcal{M})$, so $\mathcal{T}$ is
compact.
\end{proof}

When the kernel is a Mercer kernel --- symmetric and positive definite, as for
the Gaussian variant
$\kappa_{\text{gauss}}(x,y) = \exp(-\norm{Q(x)-K(y)}^2/2\sigma^2)$ --- the
spectral theorem for compact self-adjoint operators yields a decomposition
$\kappa(x,y) = \sum_{i} \lambda_i\,\phi_i(x)\phi_i(y)$ with non-negative
eigenvalues $\lambda_1\ge\lambda_2\ge\cdots\ge 0$ and orthonormal eigenfunctions
$\phi_i$, so that
\[
  \mathcal{T}f = \sum_{i=1}^{\infty}\lambda_i \inner{f}{\phi_i}_{L^2}\phi_i .
\]
The eigenfunctions are the fundamental ``attention modes'' and the eigenvalues
rank them by importance (\Cref{fig:apps:spectrum}). Because the eigenvalues
typically decay rapidly, attention acts as a natural low-pass filter: the
rank-$r$ truncation $\mathcal{T}_r$ satisfies the error bound
$\norm{\mathcal{T}f - \mathcal{T}_r f}_{L^2}\le \lambda_{r+1}\norm{f}_{L^2}$, so
a finite effective rank already captures the operator. When the kernel is
translation invariant this spectral decomposition reduces to the classical
Fourier analysis of \Cref{ch:grids}, with eigenfunctions complex exponentials
and eigenvalues the Fourier transform of the kernel --- a direct bridge between
the attention operator and the convolution.

\begin{figure}[h]
\centering
\begin{tikzpicture}[>=Stealth, font=\small, scale=0.78, every node/.style={transform shape}]
% LEFT PANEL: eigenvalue spectrum
\begin{scope}[shift={(0,0)}]
  \node[font=\bfseries, text=black] at (4.0, 6.2) {Spectrum of the attention operator};
  \fill[gdlyellow!10] (0.27, 0) rectangle (3.65, 5.2);
  \draw[->, gdlgray!70, line width=0.8pt] (-0.3, 0) -- (8.5, 0)
    node[right, font=\footnotesize, text=gdlgray!70!black] {$i$};
  \draw[->, gdlgray!70, line width=0.8pt] (0, -0.3) -- (0, 5.8)
    node[above, font=\footnotesize, text=gdlgray!70!black] {$\lambda_i$};
  \foreach \y/\lab in {1/0.2, 2/0.4, 3/0.6, 4/0.8, 5/1.0} {
    \draw[gdlgray!50] (-0.1, \y) -- (0.1, \y);
    \node[font=\tiny, text=gdlgray!70!black, anchor=east] at (-0.15, \y) {$\lab$};}
  \def\barwidth{0.45}
  \fill[gdlblue!60] (0.5-\barwidth/2, 0) rectangle (0.5+\barwidth/2, 5.0);
  \node[font=\tiny, text=gdlgray!70!black] at (0.5, -0.3) {$1$};
  \fill[gdlblue!55] (1.2-\barwidth/2, 0) rectangle (1.2+\barwidth/2, 2.5);
  \node[font=\tiny, text=gdlgray!70!black] at (1.2, -0.3) {$2$};
  \fill[gdlblue!50] (1.9-\barwidth/2, 0) rectangle (1.9+\barwidth/2, 1.48);
  \node[font=\tiny, text=gdlgray!70!black] at (1.9, -0.3) {$3$};
  \fill[gdlblue!45] (2.6-\barwidth/2, 0) rectangle (2.6+\barwidth/2, 1.0);
  \node[font=\tiny, text=gdlgray!70!black] at (2.6, -0.3) {$4$};
  \fill[gdlblue!40] (3.3-\barwidth/2, 0) rectangle (3.3+\barwidth/2, 0.72);
  \node[font=\tiny, text=gdlgray!70!black] at (3.3, -0.3) {$5$};
  \fill[gdlblue!35] (4.0-\barwidth/2, 0) rectangle (4.0+\barwidth/2, 0.55);
  \node[font=\tiny, text=gdlgray!70!black] at (4.0, -0.3) {$6$};
  \fill[gdlblue!30] (4.7-\barwidth/2, 0) rectangle (4.7+\barwidth/2, 0.43);
  \node[font=\tiny, text=gdlgray!70!black] at (4.7, -0.3) {$7$};
  \fill[gdlblue!25] (5.4-\barwidth/2, 0) rectangle (5.4+\barwidth/2, 0.35);
  \node[font=\tiny, text=gdlgray!70!black] at (5.4, -0.3) {$8$};
  \fill[gdlblue!20] (6.1-\barwidth/2, 0) rectangle (6.1+\barwidth/2, 0.28);
  \node[font=\tiny, text=gdlgray!70!black] at (6.1, -0.3) {$9$};
  \fill[gdlblue!18] (6.8-\barwidth/2, 0) rectangle (6.8+\barwidth/2, 0.24);
  \node[font=\tiny, text=gdlgray!70!black] at (6.8, -0.3) {$10$};
  \node[font=\small, text=gdlgray!70!black] at (7.6, 0.15) {$\cdots$};
  \node[font=\tiny, text=gdlgray!70!black] at (7.6, -0.3) {$n$};
  \draw[gdlred!70, line width=1.2pt, dashed, smooth]
    plot[domain=0.5:8.0, samples=50] (\x, {5.0/(\x)^1.5});
  \node[font=\scriptsize, text=gdlred!70!black, anchor=west, fill=white,
        fill opacity=0.85, text opacity=1, inner sep=2pt, rounded corners=1pt]
    at (5.5, 1.8) {$\lambda_i \sim i^{-\alpha}$};
  \draw[gdlorange!70, line width=1.0pt, densely dash dot] (3.65, -0.15) -- (3.65, 5.5);
  \node[font=\scriptsize, text=gdlorange!70!black, anchor=south, fill=white,
        fill opacity=0.85, text opacity=1, inner sep=2pt, rounded corners=1pt]
    at (3.65, 5.5) {effective rank $r$};
  \node[font=\scriptsize, text=gdlblue!70!black, align=center, anchor=north]
    at (4.0, -0.9) {$\kappa(x,y) = \displaystyle\sum_{i=1}^{\infty} \lambda_i \,\phi_i(x)\,\phi_i(y)$};
\end{scope}
% RIGHT PANEL: eigenfunctions
\begin{scope}[shift={(10.5, 0)}]
  \node[font=\bfseries, text=black] at (2.5, 6.2) {Eigenfunctions $\phi_i$};
  \node[font=\scriptsize, text=gdlblue!70!black, anchor=east] at (-0.3, 4.8) {$\phi_1$};
  \draw[->, gdlgray!50, line width=0.4pt] (0, 4.8) -- (5.2, 4.8);
  \draw[gdlgreen!70, line width=1.4pt, smooth]
    plot[domain=0:5, samples=40] (\x, {4.8 + 0.4*cos(18*\x)});
  \node[font=\tiny, text=gdlpurple!70!black, anchor=west] at (5.3, 4.8) {$\lambda_1$};
  \fill[gdlpurple!60] (5.9, 4.72) rectangle (6.9, 4.88);
  \node[font=\scriptsize, text=gdlblue!70!black, anchor=east] at (-0.3, 3.3) {$\phi_2$};
  \draw[->, gdlgray!50, line width=0.4pt] (0, 3.3) -- (5.2, 3.3);
  \draw[gdlgreen!70, line width=1.4pt, smooth]
    plot[domain=0:5, samples=60] (\x, {3.3 + 0.45*cos(72*\x)});
  \node[font=\tiny, text=gdlpurple!70!black, anchor=west] at (5.3, 3.3) {$\lambda_2$};
  \fill[gdlpurple!55] (5.9, 3.22) rectangle (6.4, 3.38);
  \node[font=\scriptsize, text=gdlblue!70!black, anchor=east] at (-0.3, 1.8) {$\phi_3$};
  \draw[->, gdlgray!50, line width=0.4pt] (0, 1.8) -- (5.2, 1.8);
  \draw[gdlgreen!70, line width=1.4pt, smooth]
    plot[domain=0:5, samples=80] (\x, {1.8 + 0.45*sin(144*\x)});
  \node[font=\tiny, text=gdlpurple!70!black, anchor=west] at (5.3, 1.8) {$\lambda_3$};
  \fill[gdlpurple!50] (5.9, 1.72) rectangle (6.2, 1.88);
  \node[font=\scriptsize, text=gdlblue!70!black, anchor=east] at (-0.3, 0.3) {$\phi_4$};
  \draw[->, gdlgray!50, line width=0.4pt] (0, 0.3) -- (5.2, 0.3);
  \draw[gdlgreen!70, line width=1.4pt, smooth]
    plot[domain=0:5, samples=100] (\x, {0.3 + 0.4*cos(216*\x)});
  \node[font=\tiny, text=gdlpurple!70!black, anchor=west] at (5.3, 0.3) {$\lambda_4$};
  \fill[gdlpurple!45] (5.9, 0.22) rectangle (6.1, 0.38);
  \node[font=\small, text=gdlgray!60] at (2.5, -0.4) {$\vdots$};
  \node[font=\scriptsize, text=gdlgreen!60!black, align=center, anchor=north]
    at (2.5, -0.8) {Increasing frequency $\longrightarrow$};
  \draw[gdlpurple!50, line width=0.6pt, decorate,
        decoration={brace, amplitude=4pt, mirror}] (7.1, 0.15) -- (7.1, 4.95);
  \node[font=\tiny, text=gdlpurple!70!black, anchor=west, align=center] at (7.4, 2.55)
    {Decay\\$\lambda_i \to 0$};
\end{scope}
\end{tikzpicture}
\caption[Spectral decomposition of attention]{Spectral decomposition of the
attention operator. \textbf{Left}: the eigenvalues $\lambda_i$ of the attention
kernel decay rapidly ($\lambda_i\sim i^{-\alpha}$), indicating a finite effective
rank $r$; the shaded region marks the dominant modes. \textbf{Right}: the
associated eigenfunctions $\phi_i$ show increasing frequency, analogous to a
Fourier basis; the purple bars indicate the magnitude of the corresponding
eigenvalue.}
\label{fig:apps:spectrum}
\end{figure}

\subsection{Attention as adaptive message passing}
\label{subsec:apps:attn-mpnn}

Read geometrically, scaled dot-product attention is message passing
(\Cref{ch:graphs}) in which the edge weights are not fixed by the domain but
\emph{computed from the data}: each element queries the others and decides how
much to listen to each. Where a convolution hard-codes the neighbourhood weights
and a graph network fixes them by the topology, attention learns them on the fly.
This places attention at the adaptive end of a spectrum whose rigid end is the
equivariant convolution --- a tension between built-in structure and learned
flexibility that recurs throughout geometric deep learning and which
\Cref{sec:apps:cnn-attention} makes precise.

\paragraph{Attention on graphs.}
The most direct geometric instance is the Graph Attention Network
(GAT)~\citep{velickovic2017graph}, which restricts attention to the neighbours
of each node and weights them adaptively. For a node $i$ with neighbourhood
$\mathcal{N}(i)$, it computes attention coefficients from the node features,
normalizes them over the neighbourhood, and aggregates,
\[
  e_{ij} = \operatorname{LeakyReLU}\!\bigl(a^\top[Wh_i \,\|\, Wh_j]\bigr),
  \qquad
  \alpha_{ij} = \softmax_j(e_{ij}),
  \qquad
  h_i' = \sigma\!\Bigl(\textstyle\sum_{j\in\mathcal{N}(i)} \alpha_{ij}\,Wh_j\Bigr).
\]
Because the coefficients depend only on the features of the nodes involved and
not on any ordering, the layer is permutation-equivariant by construction
(\Cref{fig:apps:gat}). Like all standard message-passing networks, its power to
distinguish graphs is bounded by the Weisfeiler--Leman test~\citep{xu2019how},
but the learned, anisotropic weighting of neighbours makes it markedly more
expressive in practice than fixed-weight aggregation.

\begin{figure}[h]
\centering
\begin{tikzpicture}[>=Stealth, font=\small]
\tikzset{
    targetnode/.style={circle, draw=gdlred!70, fill=gdlred!20,
        minimum size=28pt, inner sep=0pt, font=\small\bfseries, line width=1.6pt},
    neighnode/.style={circle, draw=gdlblue!60, fill=gdlblue!20,
        minimum size=22pt, inner sep=0pt, font=\small},
    nonneighnode/.style={circle, draw=gdlgray!40, fill=gdlgray!15,
        minimum size=20pt, inner sep=0pt, font=\small, text=gdlgray!70!black}
}
\node[targetnode] (vi) at (0, 0) {$v_i$};
\node[neighnode] (v1) at (-2.4, 1.8)  {$v_1$};
\node[neighnode] (v2) at (0.0, 2.6)   {$v_2$};
\node[neighnode] (v3) at (2.4, 1.4)   {$v_3$};
\node[neighnode] (v4) at (-2.2, -1.4) {$v_4$};
\node[nonneighnode] (v5) at (2.6, -1.6)  {$v_5$};
\node[nonneighnode] (v6) at (-3.6, 0.2)  {$v_6$};
\draw[gdlgray!40, thin] (v1) -- (v6);
\draw[gdlgray!40, thin] (v4) -- (v6);
\draw[gdlgray!40, thin] (v3) -- (v5);
\draw[gdlgray!40, thin] (v1) -- (v2);
\draw[gdlgray!40, thin] (vi) -- (v1);
\draw[gdlgray!40, thin] (vi) -- (v2);
\draw[gdlgray!40, thin] (vi) -- (v3);
\draw[gdlgray!40, thin] (vi) -- (v4);
\draw[gdlgray!40, thin] (v5) -- (v3);
\draw[-{Stealth[length=7pt,width=5pt]}, line width=3.2pt,
    gdlorange, opacity=0.90, shorten >=8pt, shorten <=7pt] (v2) -- (vi);
\node[font=\scriptsize, text=gdlorange!70!black, fill=white, inner sep=1.5pt,
    rounded corners=1pt] at ($(v2)!0.48!(vi) + (0.7, 0.1)$) {$\alpha_{2i} = 0.42$};
\draw[-{Stealth[length=6pt,width=4.5pt]}, line width=2.2pt,
    gdlorange, opacity=0.75, shorten >=8pt, shorten <=7pt] (v1) -- (vi);
\node[font=\scriptsize, text=gdlorange!70!black, fill=white, inner sep=1.5pt,
    rounded corners=1pt] at ($(v1)!0.48!(vi) + (-0.85, -0.15)$) {$\alpha_{1i} = 0.28$};
\draw[-{Stealth[length=5.5pt,width=4pt]}, line width=1.5pt,
    gdlorange, opacity=0.60, shorten >=8pt, shorten <=7pt] (v3) -- (vi);
\node[font=\scriptsize, text=gdlorange!70!black, fill=white, inner sep=1.5pt,
    rounded corners=1pt] at ($(v3)!0.48!(vi) + (0.85, -0.15)$) {$\alpha_{3i} = 0.19$};
\draw[-{Stealth[length=5pt,width=3.5pt]}, line width=0.9pt,
    gdlorange, opacity=0.45, shorten >=8pt, shorten <=7pt] (v4) -- (vi);
\node[font=\scriptsize, text=gdlorange!70!black, fill=white, inner sep=1.5pt,
    rounded corners=1pt] at ($(v4)!0.48!(vi) + (-0.85, 0.15)$) {$\alpha_{4i} = 0.11$};
\begin{scope}[shift={(-3.6, -2.8)}]
    \node[font=\scriptsize\bfseries, text=gdlgray!70!black, anchor=west]
        at (0, 0) {Thickness $\propto$ attention weight};
    \draw[-{Stealth[length=4pt,width=3pt]}, line width=0.9pt,
        gdlorange, opacity=0.45] (0, -0.4) -- (1.2, -0.4);
    \node[font=\tiny, text=gdlgray!70!black, anchor=west] at (1.35, -0.4) {low};
    \draw[-{Stealth[length=5pt,width=4pt]}, line width=3.2pt,
        gdlorange, opacity=0.90] (2.4, -0.4) -- (3.6, -0.4);
    \node[font=\tiny, text=gdlgray!70!black, anchor=west] at (3.75, -0.4) {high};
\end{scope}
\end{tikzpicture}
\caption[Graph attention mechanism]{The graph attention mechanism. The target
node (red) computes attention coefficients $\alpha_{ij}$ with each neighbour
(blue); the arrow thickness reflects the magnitude of the learned weight. The
network learns to assign different importance to each neighbour from the node
features alone, so the operation is invariant to how the neighbours are
ordered.}
\label{fig:apps:gat}
\end{figure}

\paragraph{Equivariant attention in 3D.}
When the inputs carry geometric coordinates, attention must respect the
symmetries of space. The general principle is clean: an attention mechanism is
$G$-equivariant precisely when the compatibility scores are $G$-invariant and
the value transformation commutes with the group action.

\begin{proposition}[Equivariant attention]
\label{prop:apps:equiv-attn}
An attention layer is $G$-equivariant if the attention scores $\alpha_{ij}$ are
invariant under the action of $G$ and the value map commutes with that action.
Then $f(g\cdot x) = g\cdot f(x)$ for every $g\in G$.
\end{proposition}

The $\SE(3)$-Transformer~\citep{fuchs2020se3transformers} realizes this for
rigid motions in three dimensions. It computes attention scores from quantities
that are invariant under rotations and translations --- scalar features,
interatomic distances $\norm{r_i - r_j}$, and inner products of vector features
--- and aggregates type-$\ell$ features so that vectors rotate with the scene
while scalars stay fixed. The Equiformer~\citep{liao2022equiformer} replaces the
dot product with an equivariant network acting on tensor products of irreducible
representations, allowing richer interactions while preserving $\SO(3)$
equivariance, and the Point Transformer~\citep{zhao2021point} brings vector
attention to point clouds. These models are the workhorses of the molecular and
physical applications of \Cref{sec:apps:molecules}.

\paragraph{Non-Euclidean and algebraic attention.}
Two further directions widen the reach of attention. Hyperbolic attention
networks~\citep{gulcehre2019hyperbolic} replace the Euclidean inner product with
operations in hyperbolic space, whose exponentially growing volume embeds
tree-like and hierarchical data with far less distortion than Euclidean space
(developed in \Cref{ch:geodesics}). Geometric Algebra Transformers
(GATr)~\citep{brehmer2023geometric} represent points, lines, and planes uniformly
as multivectors of a Clifford algebra and build $\Eucl(3)$-equivariant attention
on top, yielding a single architecture that handles diverse 3D geometric
primitives. The common thread is that the inductive bias of the geometry --- be
it negative curvature or the algebra of rigid motions --- is carried into the
attention operator itself.

\section{From convolution to attention: a unified view}
\label{sec:apps:cnn-attention}

The attention foundations of the previous section and the geometric convolutions
of \Cref{ch:grids,ch:groups,ch:graphs} are not two competing paradigms but two
points on a single continuum. This section makes the connection precise, closing
the theoretical arc of the book before turning to applications.

\begin{definition}[Generalized geometric message passing]
\label{def:apps:geom-mp}
Let $\mathcal{D}$ be a geometric domain (a Euclidean grid, a graph, a manifold,
a homogeneous space), $G$ a symmetry group, and $\mathcal{F}(\mathcal{D})$ the
space of signals on $\mathcal{D}$. A geometric message-passing operator is a
transformation $\mathcal{T}^{(G)}:\mathcal{F}(\mathcal{D})\to\mathcal{F}(\mathcal{D})$
that is $G$-equivariant ($\mathcal{T}^{(G)}(g\cdot f) = g\cdot\mathcal{T}^{(G)}(f)$),
geometrically local (its support is determined by the geometry of
$\mathcal{D}$), and compositional (composites preserve these properties).
\end{definition}

\begin{theorem}[CNN--attention correspondence]
\label{thm:apps:cnn-attn-correspondence}
There is a correspondence between geometric convolutions and geometric attention
through geometric message passing: a geometric convolution is message passing
with \emph{fixed} weights, geometric attention is message passing with
\emph{adaptive} weights, and hybrid architectures sit between, with partially
structured weights. The correspondence preserves equivariance, expressive power,
and computational complexity.
\end{theorem}

\begin{proof}[Sketch]
The equivalence holds at three levels. \emph{Algebraically}, both operations are
generalized convolutions in the group algebra $\C[G]$: for a convolution the
kernel is fixed, $k_g$; for attention it is dynamic, $\alpha_{ij}(h_i, h_j)$.
\emph{Spectrally}, in the Fourier basis of the domain both correspond to
multiplication by a transfer function --- a fixed spectral filter
$\hat{k}(\lambda)$ for convolution, an adaptive one $\hat\alpha(\lambda;h)$ for
attention (cf.\ \Cref{subsec:apps:integral-operator}). \emph{Geometrically}, the
neighbourhood structure fixes the support of integration: convolution integrates
over fixed neighbourhoods, attention integrates with content-dependent weights
that still respect the underlying geometry.
\end{proof}

\Cref{tab:apps:cnn-attn} records the correspondence concept by concept. The same
unifying principle --- operations that commute with the group action --- underlies
group convolution and $G$-equivariant attention, fixed and adaptive spectral
filters, gauge convolution and gauge-equivariant attention, and parallel
transport and Riemannian attention. The architectures of geometric deep learning
thereby form a natural hierarchy of generalization, with the rigid convolutional
family on one side and the adaptive attention family on the other, dual under
this correspondence.

\begin{table}[h]
\centering
\caption{Fundamental correspondences between geometric convolution and geometric
attention.}
\label{tab:apps:cnn-attn}
\begin{tabular}{@{}p{3.5cm}p{3.5cm}p{6cm}@{}}
\toprule
\textbf{Geometric CNN} & \textbf{Geometric attention} & \textbf{Unifying principle} \\
\midrule
Group convolution (\Cref{ch:groups}) & $G$-equivariant attention & Operations commuting with the action of $G$ \\
\addlinespace
Spectral filter $\hat{f}(\lambda) = g(\lambda)\hat{x}(\lambda)$ & Spectral attention $A = \softmax(f(\lambda_i,\lambda_j))$ & Multiplication in the Fourier domain of the Laplacian \\
\addlinespace
Gauge convolution (\Cref{ch:gauges}) & Gauge-equivariant attention & Respect for local coordinate transformations \\
\addlinespace
Steerable CNN, $f\mapsto\rho(g)f$ & Steerable attention, $A\mapsto\rho(g)A\rho(g)^{-1}$ & Coherent action on vector fields \\
\addlinespace
Spherical CNN, harmonics $Y_\ell^m$ & Spherical attention & Harmonic analysis on the sphere \\
\addlinespace
Manifold CNN, parallel transport & Riemannian attention & Intrinsic geometry of manifolds \\
\bottomrule
\end{tabular}
\end{table}

\begin{remark}[Rigidity versus adaptability]
\label{rem:apps:rigidity-adaptability}
The most fundamental distinction between the convolutional and attention
paradigms is the balance between imposed structure (rigidity) and learnable
flexibility (adaptability). One may measure the \emph{rigidity} of an
architecture by the fraction of its parameter budget devoted to imposed
symmetries and its \emph{adaptability} by the fraction devoted to
content-dependent weights, with the two competing for the same budget. The
trade-off is sharp: theoretical guarantees scale with rigidity, while expressive
capacity scales with adaptability. Along the progression from the classical CNN
(high rigidity, low adaptability) through group CNNs and graph attention to the
full transformer (low rigidity, high adaptability), this explains why
convolutional models generalize better from little data while transformers
require massive data but attain greater expressivity.
\end{remark}

\begin{remark}[Convergence of paradigms]
\label{rem:apps:convergence}
Recent architectures suggest a conceptual convergence of the two families.
Hybrid designs combine a convolutional kernel --- for local features with
geometric guarantees --- with attention heads --- for adaptive global
dependencies --- and a learnable mechanism to balance them; both families can be
approximated arbitrarily well by finite-dimensional spectral operators, and both
converge to the same limit of equivariant operators on the domain. The future of
geometric deep learning lies not in competition between paradigms but in their
synthesis: combining the stability and efficiency of geometric convolutions with
the flexibility and expressivity of adaptive attention, guided by mathematical
principles rather than heuristics. Geometry, as so often in mathematics and
physics, provides the unifying language.
\end{remark}

\section{Computer vision}
\label{sec:apps:vision}

Computer vision is the historical birthplace of the ideas in this book.
Convolutional neural networks won the ImageNet competition in
2012~\citep{krizhevsky2012alexnet} precisely because they encode a symmetry of
images: a translation of the input produces a translated feature map. As
developed in \Cref{ch:grids}, this translation equivariance is what lets a
single learned filter be reused at every position, slashing the parameter count
and providing the inductive bias that makes images learnable from finite data.
Everything that follows in computer vision can be read as an attempt to extend
this success beyond the flat pixel grid.

\paragraph{Beyond translation.}
Many visual tasks are invariant to more than translation. A galaxy, a cell
under a microscope, or an aerial image has no preferred orientation; a
classifier should not have to relearn the same pattern in every rotated copy.
Group-equivariant convolutional networks (\Cref{ch:groups}) build rotation and
reflection symmetry directly into the convolution, so that a feature detected in
one orientation is automatically detected in all of them. The payoff is sharper
on small datasets, where data augmentation cannot cover the symmetry group
densely enough and the architecture must supply the invariance instead.

\paragraph{Three-dimensional vision.}
The cleanest break from the grid comes with three-dimensional data. A LiDAR
scan or a depth sensor returns a \emph{point cloud}: an unordered set of points
in $\R^3$, with no grid, no canonical ordering, and a variable cardinality. The
relevant symmetry is twofold. First, the output must be \emph{invariant to
permutations} of the points, since the set $\{p_1,\dots,p_n\}$ does not depend on
how we list its elements. Second, for many tasks it must be invariant or
equivariant to rigid motions of the scene. The first requirement is captured
exactly by the theory of functions on sets.

\begin{theorem}[Universal representation of set functions]
\label{thm:apps:deepsets}
Any continuous permutation-invariant function $f$ of a set
$\{x_1,\dots,x_n\}\subset\R^d$ can be written in the form
\[
  f(\{x_1,\dots,x_n\}) \;=\; \rho\!\left(\sum_{i=1}^n \phi(x_i)\right),
\]
for suitable continuous functions $\phi$ and $\rho$.
\end{theorem}

This result, due to the DeepSets framework~\citep{zaheer2017deepsets}, is the
theoretical foundation of architectures such as
PointNet~\citep{qi2017pointnet}: encode each point independently with a shared
network $\phi$, pool the encodings with a permutation-invariant aggregator (a
sum or a max), and decode with $\rho$. The construction is the simplest possible
instance of the message-passing blueprint of \Cref{ch:graphs} --- a single
round of messages on a graph with no edges --- and it already guarantees the
correct invariance by design rather than by augmentation.

\paragraph{Curved domains.}
When the data lives on a curved surface rather than in free space, the gauge and
geodesic themes (\Cref{ch:geodesics,ch:gauges}) take over. Spherical signals ---
panoramic images, cosmological maps, the surface of the Earth --- are processed
by spherical convolutional networks~\citep{cohen2018sphericalcnns} that are
equivariant to the rotation group $\SO(3)$, so that rotating the sphere rotates
the response coherently. On a general mesh there is no global frame, and
gauge-equivariant networks ensure that the output does not depend on the
arbitrary local choice of coordinates. In every case the recurring lesson is the
same: identify the symmetry of the domain, then build it into the operator.

\section{Molecules and physical systems}
\label{sec:apps:molecules}

If there is one domain that vindicates geometric deep learning, it is the
physical and life sciences. Molecules \emph{are} graphs whose nodes are atoms
and whose edges are bonds, embedded in $\R^3$; the laws governing them are
manifestly invariant under rigid motions of space. A model that ignores these
symmetries wastes capacity relearning physics that could simply be assumed.

\paragraph{Molecular property prediction.}
Graph neural networks predict pharmacological and quantum-chemical properties
directly from the molecular graph, replacing the hand-crafted descriptors and
string encodings (such as SMILES) of classical cheminformatics, which discard
the three-dimensional information that governs biological activity. When the
target depends on the molecular conformation --- energies, forces, dipole
moments --- the model must be $\Eucl(3)$- or $\SE(3)$-equivariant: rotating the
molecule must rotate predicted vectors and leave predicted scalars unchanged.
The equivariant attention architectures of \Cref{sec:apps:attn-mpnn}, together
with tensor-field and equivariant message-passing networks~\citep{satorras2021enequivariant},
are precisely the tools that enforce this.

\begin{example}[AlphaFold and protein structure]
\label{ex:apps:alphafold}
The most celebrated success of the field is
AlphaFold~\citep{jumper2021alphafold}, which solved the decades-old problem of
predicting a protein's three-dimensional structure from its amino-acid sequence.
At its core lies an attention-based architecture that reasons about the geometry
of residues in space while respecting the $\SE(3)$ symmetry of rigid motions:
the predicted structure transforms consistently when the frame of reference is
rotated or translated. Geometric equivariance is not a stylistic choice here but
a structural requirement, and it is a large part of why the model generalizes so
well. AlphaFold is, in this sense, the paradigmatic illustration of the
theory--architecture connection that this book traces: the abstract $\SE(3)$
equivariance of \Cref{ch:groups} made concrete in a system that reshaped
structural biology.
\end{example}

\paragraph{Physical simulation.}
The same principles extend from static structures to dynamics. Equivariant
networks learn to simulate physical systems --- fluids, deformable materials,
many-body dynamics --- by operating on irregular meshes or particle sets while
respecting the symmetries of the underlying differential equations. Because the
symmetry is built in, the predictions are physically consistent by construction
(a rotated input yields a correspondingly rotated forecast) and the models often
match or exceed classical numerical solvers at a fraction of the cost. A single
foundation model can even be pretrained across small molecules and proteins at
once: equivariance is an architectural property, independent of the data, so a
network trained to be equivariant on one domain remains equivariant when
transferred to another with the same symmetry~\citep{jiao2024ept}.

\paragraph{Materials science.}
Crystalline solids add a further layer of symmetry: the discrete translational
and point-group symmetries of the crystal lattice. Encoding these symmetries in
equivariant networks lets models predict mechanical, thermal, and electronic
properties of candidate compounds, accelerating the search for new materials. As
in the molecular case, the geometric prior turns scarce, expensive labels into a
tractable learning problem.

\section{Recommender systems and social networks}
\label{sec:apps:graphs}

Relational data --- who interacts with whom, who buys what --- is graph-shaped
by nature, and here the governing symmetry is permutation invariance: the
identity of a user does not depend on the order in which we list the accounts.
Graph neural networks (\Cref{ch:graphs}) exploit this directly, propagating
information along the edges of the interaction graph so that each node's
representation is shaped by its neighbourhood.

\paragraph{Recommendation and social analysis.}
Recommender systems model users and items as nodes of a bipartite graph and
learn embeddings by message passing over observed interactions, capturing
collaborative signals that flat feature tables miss. Social-network analysis
uses the same machinery for node classification (inferring user attributes),
link prediction (suggesting connections), and community detection. In all of
these the irregular topology --- nodes with wildly different numbers of
neighbours and no natural ordering --- is exactly what convolutional models
cannot handle and graph networks are built for.

\paragraph{Hierarchies and hyperbolic geometry.}
Many relational datasets are not merely irregular but \emph{hierarchical}:
taxonomies, organizational charts, knowledge graphs, and the follower structure
of social networks all branch like trees. The number of nodes at distance $r$
from a root grows exponentially in $r$, and Euclidean space simply does not have
enough room to embed such growth without crushing distances. Hyperbolic space
does: its volume also grows exponentially, so a tree embeds with almost no
distortion (\Cref{fig:apps:hyperbolic}). This is the geometric motivation behind
hyperbolic embeddings and the hyperbolic attention networks of
\Cref{sec:apps:attn-mpnn}, which place knowledge-graph entities and hierarchical
users in negatively curved space.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    root node/.style={circle, draw=gdlpurple!80, fill=gdlpurple!70,
        minimum size=12pt, inner sep=0pt},
    depth1 node/.style={circle, draw=gdlblue!80, fill=gdlblue!60,
        minimum size=10pt, inner sep=0pt},
    depth2 node/.style={circle, draw=gdlgreen!80, fill=gdlgreen!60,
        minimum size=10pt, inner sep=0pt},
    leaf node/.style={circle, draw=gdlyellow!80, fill=gdlyellow!70,
        minimum size=8pt, inner sep=0pt},
    tree edge/.style={draw=gdlgray!50, line width=0.8pt},
    scale=0.82, every node/.style={transform shape}
]
% LEFT: Euclidean
\begin{scope}[shift={(0,0)}]
    \coordinate (r) at (0, 1.9);
    \coordinate (d1-0) at (-1.8, 0.4); \coordinate (d1-1) at ( 1.8, 0.4);
    \coordinate (d2-0) at (-2.7, -1.1); \coordinate (d2-1) at (-0.9, -1.1);
    \coordinate (d2-2) at ( 0.9, -1.1); \coordinate (d2-3) at ( 2.7, -1.1);
    \coordinate (d3-0) at (-3.15, -2.6); \coordinate (d3-1) at (-2.25, -2.6);
    \coordinate (d3-2) at (-1.35, -2.6); \coordinate (d3-3) at (-0.45, -2.6);
    \coordinate (d3-4) at ( 0.45, -2.6); \coordinate (d3-5) at ( 1.35, -2.6);
    \coordinate (d3-6) at ( 2.25, -2.6); \coordinate (d3-7) at ( 3.15, -2.6);
    \draw[dashed, gdlgray!50, line width=0.8pt] (0,-0.35) circle (3.6);
    \draw[tree edge] (r) -- (d1-0); \draw[tree edge] (r) -- (d1-1);
    \draw[tree edge] (d1-0) -- (d2-0); \draw[tree edge] (d1-0) -- (d2-1);
    \draw[tree edge] (d1-1) -- (d2-2); \draw[tree edge] (d1-1) -- (d2-3);
    \draw[tree edge] (d2-0) -- (d3-0); \draw[tree edge] (d2-0) -- (d3-1);
    \draw[tree edge] (d2-1) -- (d3-2); \draw[tree edge] (d2-1) -- (d3-3);
    \draw[tree edge] (d2-2) -- (d3-4); \draw[tree edge] (d2-2) -- (d3-5);
    \draw[tree edge] (d2-3) -- (d3-6); \draw[tree edge] (d2-3) -- (d3-7);
    \node[root node] at (r) {};
    \node[depth1 node] at (d1-0) {}; \node[depth1 node] at (d1-1) {};
    \foreach \i in {0,...,3} {\node[depth2 node] at (d2-\i) {};}
    \foreach \i in {0,...,7} {\node[leaf node] at (d3-\i) {};}
    \node[font=\footnotesize\itshape, text=gdlgray!70!black] at (0, -4.6)
        {Euclidean embedding};
\end{scope}
% RIGHT: Poincaré disk
\begin{scope}[shift={(9.0, -0.35)}]
    \pgfmathsetmacro{\R}{3.6}
    \foreach \rad in {0.30, 0.52, 0.70, 0.83, 0.92} {
        \draw[gdlgray!20, thin] (0,0) circle (\rad*\R);}
    \foreach \ang in {0,30,60,90,120,150,180,210,240,270,300,330} {
        \draw[gdlgray!20, thin] (0,0) -- ({\ang}:\R);}
    \draw[black, line width=1.2pt] (0,0) circle (\R);
    \pgfmathsetmacro{\rone}{0.6044*\R}
    \pgfmathsetmacro{\rtwo}{0.8854*\R}
    \pgfmathsetmacro{\rthree}{0.9505*\R}
    \coordinate (pr) at (0, 0);
    \coordinate (pd1-0) at ({90}:{\rone}); \coordinate (pd1-1) at ({270}:{\rone});
    \coordinate (pd2-0) at ({50}:{\rtwo}); \coordinate (pd2-1) at ({130}:{\rtwo});
    \coordinate (pd2-2) at ({230}:{\rtwo}); \coordinate (pd2-3) at ({310}:{\rtwo});
    \coordinate (pd3-0) at ({35}:{\rthree}); \coordinate (pd3-1) at ({65}:{\rthree});
    \coordinate (pd3-2) at ({115}:{\rthree}); \coordinate (pd3-3) at ({145}:{\rthree});
    \coordinate (pd3-4) at ({215}:{\rthree}); \coordinate (pd3-5) at ({245}:{\rthree});
    \coordinate (pd3-6) at ({295}:{\rthree}); \coordinate (pd3-7) at ({325}:{\rthree});
    \draw[tree edge] (pr) -- (pd1-0); \draw[tree edge] (pr) -- (pd1-1);
    \draw[tree edge] (pd1-0) -- (pd2-0); \draw[tree edge] (pd1-0) -- (pd2-1);
    \draw[tree edge] (pd1-1) -- (pd2-2); \draw[tree edge] (pd1-1) -- (pd2-3);
    \draw[tree edge] (pd2-0) -- (pd3-0); \draw[tree edge] (pd2-0) -- (pd3-1);
    \draw[tree edge] (pd2-1) -- (pd3-2); \draw[tree edge] (pd2-1) -- (pd3-3);
    \draw[tree edge] (pd2-2) -- (pd3-4); \draw[tree edge] (pd2-2) -- (pd3-5);
    \draw[tree edge] (pd2-3) -- (pd3-6); \draw[tree edge] (pd2-3) -- (pd3-7);
    \node[root node] at (pr) {};
    \node[depth1 node] at (pd1-0) {}; \node[depth1 node] at (pd1-1) {};
    \foreach \i in {0,...,3} {\node[depth2 node] at (pd2-\i) {};}
    \foreach \i in {0,...,7} {\node[leaf node] at (pd3-\i) {};}
    \node[font=\footnotesize\itshape, text=gdlgray!70!black] at (0, -4.25)
        {Poincar\'e-disk embedding};
\end{scope}
\end{tikzpicture}
\caption[Euclidean versus hyperbolic embedding of a tree]{Embedding a balanced
tree in Euclidean space (left) and in the Poincar\'e disk (right). In the plane,
successive levels crowd together and sibling subtrees overlap, distorting
distances; in hyperbolic space the exponentially growing volume gives every
level room to spread out, so the tree embeds with little distortion. This is why
hierarchical relational data is naturally modelled in negatively curved space.}
\label{fig:apps:hyperbolic}
\end{figure}

\paragraph{Temporal graphs: fraud and risk.}
Financial transactions form graphs whose topology evolves in time. Graph neural
networks operating on these temporal graphs detect fraud rings and systemic
risk by recognizing structural patterns --- unusual subgraphs, sudden changes in
connectivity --- that tabular models, blind to the relational structure, cannot
see. The geometric prior is again decisive: it is the \emph{shape} of the
interaction pattern, not the individual features, that carries the signal.

\section{A common principle}
\label{sec:apps:common}

Across these domains a single message recurs. In computer vision the symmetry is
translation, rotation, or the gauge freedom of a curved surface; in molecular
and physical modelling it is the Euclidean group of rigid motions; in relational
data it is permutation of nodes, and the geometry of the embedding space may be
flat or negatively curved. In every case the same recipe applies: identify the
symmetry the problem respects, build it into the architecture, and let the
network spend its capacity on what actually varies.

\begin{remark}[Geometry as the organizing principle]
\label{rem:apps:unification}
The breadth of these applications is itself the strongest evidence for the
central claim of this book. Convolutional networks for images, equivariant
transformers for proteins, and graph networks for social data look like
unrelated tools, yet each is an instance of equivariant message passing,
differing only in the domain and the symmetry group it respects --- and, by
\Cref{thm:apps:cnn-attn-correspondence}, differing only in whether the messages
carry fixed or adaptive weights. Geometry is not a domain-specific trick but the
common principle that geometric deep learning identifies, formalizes, and
exploits.
\end{remark}

Beyond these consolidated areas, the same ideas are opening lines of research
wherever geometric structure is fundamental: computational neuroscience, where
graph networks model structural and functional brain connectivity and reveal
patterns of neurological disorder invisible to tabular methods; and climate
science, where the spherical geometry of the Earth invites $\SO(3)$-equivariant
spherical networks for weather and climate prediction. The diversity of these
settings underscores that respecting the geometry of data is not an
optimization detail but the organizing principle of the field, and the bridge
from the theory of the preceding chapters to the systems that are reshaping
science and technology.
